---
title: "LK1 pilot run"
subtitle: "`r params$run_id`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  run_id: "260423_VH01624_453_222HWMYNX"
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  sample_short: "LK1"
  sample_name: "LK1-GEX"
---

<!--
# ------------------------------------------------------------------
# LK1 pilot run - step 18 of 18
#
# Renders the lab archive report for the pilot run from the outputs of
# steps 01-17. Everything that can be read off disk is read off disk, so
# the report cannot drift from the results: tables come from the summary
# CSVs each step wrote, figures are rasterised straight out of the PDFs
# those steps produced, and the timeline is built from file modification
# times.
#
# Missing inputs are skipped silently rather than erroring, so a partial
# render is the expected case, not a failure.
#
# Same output format as 21.run_qc_report.Rmd (html_document, flatly,
# floating toc, folded code) and it reuses that script's embed_pdf_figure()
# so PDF figures render inline rather than being named as paths.
#
# Render with:
#   Rscript -e 'rmarkdown::render("18.LK1_pilot_run_report.Rmd")'
# pandoc is not on PATH on this laptop; if rendering outside RStudio, point
# at the one inside the RStudio bundle first:
#   export RSTUDIO_PANDOC=/Applications/RStudio.app/Contents/Resources/app/quarto/bin/tools/aarch64
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
  library(stringr)
  library(purrr)
  library(knitr)
})

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

# ---------------------------------------------------------------
# Convert a pipeline-generated PDF figure to PNG and return a markdown image
# tag, so figures actually render inline in the HTML report instead of just
# being named as a file path. Falls back to a plain path link if pdftools
# isn't installed or conversion fails for any reason.
# (copied verbatim from 21.run_qc_report.Rmd so the two reports behave the same)
# ---------------------------------------------------------------
embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) {
    cat(sprintf("\n_(pdftools not installed - showing path instead)_ `%s`\n\n", pdf_path))
    return(invisible(NULL))
  }
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE)
    TRUE
  }, error = function(e) {
    cat(sprintf("\n_(failed to render `%s`: %s)_\n\n", basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

# Raster images (the copykat heatmaps are jpeg) go in directly - pandoc
# base64-embeds them when self_contained is on, which is the html_document default.
embed_image <- function(path, width_pct = 100) {
  if (is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', path, width_pct))
}

# One figure with its caption and description, then the image.
# Anything not on disk is skipped without comment.
show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  if (grepl("\\.pdf$", path, ignore.case = TRUE)) {
    embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
  } else {
    embed_image(path, width_pct = width_pct)
  }
}

`%||%` <- function(a, b) if (is.null(a)) b else a

proj    <- params$proj
run     <- params$run_id
res     <- file.path(proj, "results")
raw_dir <- file.path(proj, "raw", run)

# ---------------------------------------------------------------
# Small helpers - all reads are defensive: missing files return NULL
# rather than erroring, so the report renders even for a partial run.
# ---------------------------------------------------------------
read_if <- function(path, ...) {
  if (!file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)),
             silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n")
  print(kable(df, caption = caption))
  cat("\n\n")
}

rd <- function(...) file.path(res, ..., run)

# All output files for this run, with dates - drives the timeline.
run_files <- function() {
  dirs <- list.dirs(res, recursive = FALSE, full.names = TRUE)
  hits <- character(0)
  for (d in dirs) {
    p <- file.path(d, run)
    if (dir.exists(p)) hits <- c(hits, list.files(p, recursive = TRUE, full.names = TRUE))
  }
  if (!length(hits)) return(NULL)
  tibble(
    path = hits,
    rel  = sub(paste0("^", res, "/"), "", hits),
    date = as.Date(file.info(hits)$mtime),
    ext  = tolower(tools::file_ext(hits))
  ) %>% filter(!grepl("(^|/)\\._", rel))
}

files_tbl <- run_files()

# Find one file by the tail of its path, so the figure sections do not have to
# hardcode which results directory a figure ended up in.
fig <- function(pattern) {
  if (is.null(files_tbl)) return(NA_character_)
  hit <- files_tbl$path[grepl(pattern, files_tbl$rel, fixed = TRUE)]
  if (!length(hit)) NA_character_ else hit[1]
}
```

**Run:** `` `r run` ``

**Status:** superseded - LK1 was re-sequenced as `260522_VH01624_461_222JLJVNX`. Everything
below is the first pass on the pilot data and should not be quoted as a current result.

**Generated:** `r Sys.Date()` by `18.LK1_pilot_run_report.Rmd`, from the outputs of steps
01-17 in this folder.

This is the first run of the project and most of the pipeline was worked out on it. Several
of the scripts that produced these outputs have since been renumbered or replaced; the copies
in this folder are the versions that ran, renumbered into execution order.

# 1. Run, libraries and samples

```{r read-counts, results='asis'}
rc <- read_if(file.path(raw_dir, "fastq_read_counts.tsv"),
              delim = "\t", col_names = c("fastq", "reads"))
if (!is.null(rc)) {
  rc <- rc %>%
    mutate(library = case_when(
      grepl("GEX", fastq)  ~ "GEX",
      grepl("-ADT", fastq) ~ "ADT (TotalSeq-A)",
      grepl("-HTO", fastq) ~ "HTO (TotalSeq-A)",
      TRUE                 ~ "other")) %>%
    arrange(desc(reads)) %>%
    mutate(reads = format(reads, big.mark = ","))
}
show_table(rc, caption = "R1 read counts per library (01.count_reads.sh)")
```

Read counts are line count / 4 on R1. The HTO library is far shallower than ADT, which is the
expected ratio but worth remembering when reading the demux figures.

Four samples were hashed into the well. The map is hardcoded in
`02.initial_QC_and_HTO_demux.R`:

| HTO barcode | Sample |
|---|---|
| `HTO1-GTCAACTCTTTAGCG` | MOLM13 (AML cell line) |
| `HTO2-TGATGGCCTATTGGG` | HBDN206-MNpCT |
| `HTO3-TTCCGCCTCTCTTTG` | HBDN392-AML-MDS |
| `HTO4-AGTAAGTTCAGCGTA` | HBDN501-AML-KMT2A |

Note the naming drift: the HTO map writes them with underscores (`HBDN206_MNpCT`), the
figures and later metadata use hyphens (`HBDN206-MNpCT`). Same samples.

```{r demux-tables, results='asis'}
show_table(read_if(file.path(rd("seurat_demux"), "demux_summary.csv")),
           caption = "HTO demultiplexing outcome")
show_table(read_if(file.path(rd("seurat_demux"), "sample_qc_summary.csv")),
           caption = "Per-sample QC medians")
```

# 2. Data locations

| What | Location |
|---|---|
| Sequencing facility output | `` `r file.path("/pipeline/Runs/NextSeq", run, "ProjectFolders/Project_Lev-Kats")` `` |
| Working copy, cluster | `` `r file.path("/scratch/users/nbartonicek/projects/amgen/raw", run)` `` - `Sample_LK1-GEX/`, `Sample_LK1-TotalSeqA-ADT/`, `Sample_LK1-TotalSeqA-HTO/` |
| Reference | Cell Ranger annotation `refdata-gex-GRCh38-2024-A`, under `annotation/` |

```{r output-dirs, results='asis'}
if (!is.null(files_tbl)) {
  files_tbl %>%
    mutate(dir = sub("/.*", "", rel)) %>%
    group_by(dir) %>%
    summarise(first = min(date), last = max(date), files = n(), .groups = "drop") %>%
    arrange(first) %>%
    show_table(caption = paste0("Output directories under results/&lt;dir&gt;/", run, "/"))
}
```

# 3. Timeline

```{r timeline, results='asis'}
if (!is.null(files_tbl)) {
  files_tbl %>%
    mutate(dir = sub("/.*", "", rel)) %>%
    group_by(date) %>%
    summarise(directories = paste(sort(unique(dir)), collapse = ", "),
              files = n(), .groups = "drop") %>%
    arrange(date) %>%
    show_table(caption = "What was written, by day")
}
```

# 4. What was run, in order

The scripts in this folder are numbered in execution order. Steps 03 and 04 were later
re-pointed at the re-sequenced run, so their frozen copies carry `run = 260522` - set it back
to the pilot run to reproduce. Their pilot outputs are dated 2026-05-01 to 05-04, three weeks
before 260522 existed.

| Step | Script | What it does |
|---|---|---|
| - | *(not preserved)* | Cell Ranger count, `~/tools/cellranger-10.0.0/cellranger`. Step 02 reads its `filtered_feature_bc_matrix`. |
| 01 | `01.count_reads.sh` | R1 read counts per library. |
| 02 | `02.initial_QC_and_HTO_demux.R` | Seurat object, HTO + ADT attached, CLR, `HTODemux(positive.quantile = 0.80)`. |
| 03 | `03.emptydrops.R` | emptyDrops on the raw matrix, FDR <= 0.01. Produces the barcode whitelist. |
| 04 | `04.cellSNP_vireo.sh` | cellsnp-lite pileup over those barcodes, then vireo with `N_DONORS=4`. |
| 05 | `05.demux_vireo_and_doublets.R` | Vireo calls onto the object, HTO cross-tab, doublet detection. |
| 06 | `06.CITE_qc.R` | CITE QC on emptyDrops cells: raw ADT, CLR, marker distributions, overlays. |
| 07 | `07.evaluate_DSB.R` | DSB using raw background droplets, with IgG-control subsampling. |
| 08 | `08.combine_cite.R` | Folds CITE assays into one object, writes the `analysis_bundle`. |
| 09 | `09.annotate_reference_setup.R` | BoneMarrowMap Symphony reference, uwot path, first projection. |
| 10 | `10.annotate_ADT_CLR.R` | Projection on the ADT/CLR object. |
| 11 | `11.annotate_CITE_DSB.R` | Same projection on the CITE_DSB object. |
| 12 | `12.compare_DSB_vs_CLR_one_sample.R` | DSB vs CLR in one sample (HBDN206-MNpCT). |
| 13 | `13.compare_DSB_vs_CLR_all_samples.R` | DSB vs CLR across all samples. Writes `data_integration/` 01-06. |
| 14 | `14.data_integration_composition.R` | Composition by sample, DSB protein-space UMAP, CopyKAT against annotation. Writes `data_integration/` 07-11. |
| 15 | `15.copykat_pooled.R` | CopyKAT, all cells pooled. |
| 16 | `16.copykat_per_sample.R` | CopyKAT per donor. |
| 17 | `17.surfaceome_trial.R` | First surfaceome pass. Exploratory, saves nothing. |
| 18 | `18.LK1_pilot_run_report.Rmd` | This report. |

Key parameters in step 02: `CreateSeuratObject(min.cells = 3, min.features = 50)`,
`percent.mt` from `^MT-`, HTO and ADT normalised CLR with `margin = 2`,
`HTODemux(positive.quantile = 0.80)`, RNA scaled regressing `percent.mt`, PCA and UMAP on
`dims = 1:30`, `FindClusters(resolution = 0.5)`.

The 0.80 positive quantile is the value that later got swept - `demux_comparison/` carries
`_all_positive_quantiles` and `_positive_quantile_0.99` variants of every figure.

# 5. Results

## 5.1 First-pass QC and HTO demultiplexing

```{r figs-qc, results='asis'}
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/01_initial_qc_violin.pdf"),
  "nFeature_RNA, nCount_RNA and percent.mt by `hash.ID`. First look - checks no hashed sample is degraded relative to the others.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/02_hto_ridgeplots.pdf"),
  "CLR-normalised HTO signal per hashtag. This is the figure that tells you whether hashing worked at all; clean bimodality per HTO is what you want.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/03_umap_demux.pdf"),
  "HTO global classification, assigned `hash.ID`, and top HTO by raw count. The third panel is the sanity check on the second.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/05_hto_featureplots.pdf"),
  "Each HTO on the UMAP. Confirms the hashed samples occupy distinct territory rather than smearing.")
```

## 5.2 Genotype demultiplexing: HTO vs vireo

```{r demux-comparison-tables, results='asis'}
dc <- file.path(rd("demux_comparison"), params$sample_name)
show_table(read_if(file.path(dc, "demux_efficiency_summary.csv")), caption = "Demux efficiency, HTO vs vireo")
show_table(read_if(file.path(dc, "hto_vireo_agreement_summary.csv")), caption = "HTO / vireo agreement")
```

```{r figs-demux, results='asis'}
show_fig(fig("demux_comparison/260423_VH01624_453_222HWMYNX/LK1-GEX/01_demux_efficiency_barplot.pdf"),
  "Singlet / doublet / negative rates for HTO and vireo side by side. The headline comparison.")
show_fig(fig("demux_comparison/260423_VH01624_453_222HWMYNX/LK1-GEX/04_upset_HTO_vireo_overlap.pdf"),
  "UpSet of the call overlap.")
show_fig(fig("genotype_demux/260423_VH01624_453_222HWMYNX/LK1-GEX/vireo/fig_GT_distance_estimated.pdf"),
  "Vireo estimated genotype distance between donors. Confirms the donors are genetically distinguishable in the first place.")
```

## 5.3 ADT / CITE-seq QC

```{r figs-adt, results='asis'}
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/06_ADT_QC.pdf"),
  "ADT totals and features per cell.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/07_top_ADT_markers.pdf"),
  "Top 5 ADT markers by mean CLR plus biotin. Biotin was added deliberately here - this is the start of the biotin thread that runs through to July.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/08_top_ADT_cluster_markers_dotplot.pdf"),
  "Top ADT markers per RNA cluster.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/10_top10_discriminative_ADT_plus_biotin_UMAP.pdf"),
  "Most discriminative ADTs with biotin overlaid.")
show_fig(fig("seurat_demux/260423_VH01624_453_222HWMYNX/11_biotin_expression_by_cluster_violin.pdf"),
  "Biotin by cluster - the first look at the heterogeneity later split high vs low.")
```

## 5.4 Annotation (BoneMarrowMap)

```{r figs-annot, results='asis'}
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/projectionFigures/01_mapping_error_QC.pdf"),
  "BoneMarrowMap mapping error. This is the QC that decides which cells get an annotation at all.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/projectionFigures/02_projected_predicted_celltypes_pass_only.pdf"),
  "Predicted cell types on the reference UMAP, cells passing mapping QC only.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/projectionFigures/03_projected_samples_pass_only.pdf"),
  "Samples on the projected UMAP.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/projectionFigures/04_projected_pseudotime_pass_only.pdf"),
  "Projected pseudotime.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/projectionFigures/05_projection_by_sampleID_BMM_helper_manual_split.pdf"),
  "Per-sample projection using the BoneMarrowMap helper.")
```

```{r figs-composition, results='asis'}
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/08_celltype_composition_by_sampleID_stacked_bar.pdf"),
  "Broad cell type composition per sample.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/11_lineage_composition_by_sampleID.pdf"),
  "Lineage composition per sample.")
show_fig(fig("seurat_annotated/260423_VH01624_453_222HWMYNX/13_specific_celltype_composition_by_sampleID.pdf"),
  "Fine-grained cell type composition per sample.")
```

## 5.5 DSB vs CLR

`06_ADT_CLR_top_marker_projection.pdf` and its DSB counterpart were built from independently
ranked marker lists, so the two panels do not show the same proteins and are not directly
comparable. Only the CLR panel is shown; the quantitative comparison is the mean-shift and
specificity work below.

```{r figs-dsb-clr, results='asis'}
show_fig(fig("projectionFigures/06_ADT_CLR_top_marker_projection.pdf"),
  "Top markers under CLR, on the projected UMAP.")
show_fig(fig("DSB_vs_CLR_HBDN206-MNpCT/01_marker_mean_shift_DSB_minus_CLR.pdf"),
  "Per-marker mean shift, DSB minus CLR, for HBDN206-MNpCT (step 12).")
show_fig(fig("data_integration/260423_VH01624_453_222HWMYNX/01_background_suppression.pdf"),
  "How much background DSB removes, measured on the isotype / background markers. The main argument for using DSB.")
show_fig(fig("data_integration/260423_VH01624_453_222HWMYNX/02_expected_marker_specificity.pdf"),
  paste("Whether expected markers land on the expected cell types, scored against a",
        "hand-curated panel defined in step 13: CD34 / CD117 / CD133 for stem-progenitor,",
        "CD33 / CD14 / CD163 / CD11b for myeloid-DC, CD4 / CD8 / CD19 / CD20 / BCMA /",
        "Syndecan-1 for lymphoid, and CD71 for erythroid. For each marker the script takes",
        "the **mean signal in its expected broad cell type minus the median across all",
        "other broad types**, computes that separately under CLR and under DSB, and plots",
        "one against the other with the diagonal drawn. Points above the diagonal are",
        "markers DSB made more specific. The underlying numbers, including the per-marker",
        "delta, are in `02_expected_marker_specificity.csv`."))
show_fig(fig("data_integration/260423_VH01624_453_222HWMYNX/03_marker_leakage_diffusion.pdf"),
  paste("Marker signal outside the population it belongs to, on the same expected-marker",
        "panel. Per marker the script computes `inside_mean` (mean signal in the expected",
        "population), `outside_mean` (mean everywhere else), the percentage of outside",
        "cells scoring above zero, and **`leakage_ratio = outside_mean / inside_mean`**.",
        "Lower is better: a genuinely specific marker should carry little signal outside",
        "its own population. `03_marker_leakage_diffusion.csv` is sorted by",
        "`delta_leakage_ratio_DSB_minus_CLR`, so the markers DSB cleans up most are at",
        "the top."))
```

```{r dsb-metrics, results='asis'}
di <- rd("data_integration")
show_table(read_if(file.path(di, "06_summary_metrics.csv")), caption = "Summary metrics, DSB vs CLR")
show_table(read_if(file.path(di, "01_background_suppression.csv")), caption = "Background suppression", n = 30)
show_table(read_if(file.path(di, "02_expected_marker_specificity.csv")), caption = "Expected marker specificity", n = 30)
show_table(read_if(file.path(di, "03_marker_leakage_diffusion.csv")), caption = "Marker leakage", n = 30)
```

## 5.6 CNV calls (CopyKAT)

CopyKAT was run twice: once on all cells pooled (step 15) and then per donor (step 16). The
per-donor run is what produced the output below.

```{r figs-cnv, results='asis'}
ck <- file.path(rd("seurat_annotated"), "copykat_annotated")
if (dir.exists(ck)) {
  # only subdirectories that actually hold copykat output - the folder also carries
  # scratch directories that would otherwise render as empty sections
  donors <- setdiff(list.dirs(ck, recursive = FALSE, full.names = FALSE), "")
  donors <- donors[vapply(donors, function(d)
    any(file.exists(file.path(ck, d, c("01_copykat_call_composition_by_sampleID.pdf",
                                       "copykat_call_composition_by_sampleID.csv",
                                       "LK1_copykat_copykat_heatmap.jpeg")))), logical(1))]
  for (d in donors) {
    cat("\n### ", d, "\n\n", sep = "")
    show_table(read_if(file.path(ck, d, "copykat_call_composition_by_sampleID.csv")),
               caption = paste("CopyKAT calls -", d))
    show_fig(file.path(ck, d, "01_copykat_call_composition_by_sampleID.pdf"),
             "Aneuploid / diploid proportions.", width_pct = 70)
    show_fig(file.path(ck, d, "02_copykat_calls_on_projected_RNA_UMAP.pdf"),
             "CopyKAT calls on the projected UMAP.")
    show_fig(file.path(ck, d, "03_copykat_by_BoneMarrowMap_broad_celltype.pdf"),
             "Calls against annotation. The check that 'aneuploid' is not just landing on one cell type for technical reasons.")
    show_fig(file.path(ck, d, "LK1_copykat_copykat_heatmap.jpeg"),
             "The CopyKAT CNV heatmap.")
  }
}
```

# 6. Tools

| Tool | Version | Source |
|---|---|---|
| Cell Ranger | 10.0.0 | `~/tools/cellranger-10.0.0/cellranger` |
| Cell Ranger annotation | `refdata-gex-GRCh38-2024-A` | `annotation/` |
| CITE-seq-Count | to confirm | `scripts/3.hto_count.sbatch`, `4.cite_seq_count.sbatch` |
| cellsnp-lite | to confirm | `04.cellSNP_vireo.sh` |
| vireo | to confirm | `04.cellSNP_vireo.sh` |

R side, the session that renders this report:

```{r session}
sessionInfo()
```

# 7. Caveats

1. **The run is superseded.** LK1 was re-sequenced as `260522_VH01624_461_222JLJVNX`, and the
   re-sequenced data is what later analyses use.

```{r drops, results='asis'}
cat("\n**Scripts dropped when collating this folder**, and why:\n\n")
cat("| Dropped | Reason |\n|---|---|\n")
cat("| `backup/8.create_seurat_object.R` | Byte-identical to `scripts/6a.initial_QC.R` (step 02). |\n")
cat("| `backup/10b.annotate_DSB.R` | Byte-identical to `scripts/12.identify_cancer_cells.R` (step 11). |\n")
cat("| `backup/4.initial_QC.R` | Earlier, shorter version of step 02. |\n")
cat("| `backup/11a.example_cite_copykat.R` | Exploratory scratch, not part of the pipeline. |\n")
cat("| `add_cite_for_hiep.R` | One-off to hand CITE data to a colleague, no analytical result. |\n\n")
```

Note that `scripts/12.identify_cancer_cells.R` does not identify cancer cells - it is the
BoneMarrowMap projection on the DSB object, byte-identical to `backup/10b.annotate_DSB.R`.
The name is misleading, so it is `11.annotate_CITE_DSB.R` here.
