---
title: "Benchmarking: demultiplexing with genotyping vs HTO"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  pilot_run: "260423_VH01624_453_222HWMYNX"
  runs: !r c("260522_VH01624_461_222JLJVNX","260528_VH01624_464_222K7VKNX","260717_VH01624_477_222KG22NX")
---

<!--
# ------------------------------------------------------------------
# Demultiplexing benchmark: genotyping vs HTO - step 07 of 07
#
# Renders the lab archive report from the outputs of steps 01-06. Tables come
# from the summary CSVs each step wrote, figures are rasterised out of the PDFs.
#
# Render with ./render_report.sh, which also prints the HTML to PDF.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) FALSE)
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

read_if <- function(path, ...) {
  if (!file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj  <- params$proj
res   <- file.path(proj, "results")
nf    <- file.path(proj, "results_nf")
pilot <- params$pilot_run
runs  <- params$runs
short <- function(r) sub("_VH.*", "", r)

# The two sources lay their output out differently: the standalone runs are flat
# under demux_comparison/<run>/<sample>/, while the pipeline nests them under
# results_nf/<run>/06_demux_comparison/<sample>/figures/ and .../tables/. Both
# lookups below therefore recurse and match on file name.
dc_root <- function(run, source = c("standalone", "pipeline")) {
  source <- match.arg(source)
  p <- if (source == "standalone") file.path(res, "demux_comparison", run)
       else file.path(nf, run, "06_demux_comparison")
  if (dir.exists(p)) p else NA_character_
}

# find one file by exact name anywhere under a comparison root
dc_file <- function(root, name) {
  if (is.na(root)) return(NA_character_)
  hits <- list.files(root, pattern = paste0("^", gsub("\\.", "\\\\.", name), "$"),
                     recursive = TRUE, full.names = TRUE)
  hits <- hits[!grepl("(^|/)\\._", hits)]
  if (!length(hits)) NA_character_ else hits[1]
}

dc_pdfs <- function(root) {
  if (is.na(root)) return(character(0))
  f <- list.files(root, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE)
  f[!grepl("(^|/)\\._", f)]
}

FIGS <- unique(c(
  unlist(lapply(c(pilot, runs), function(r) dc_pdfs(dc_root(r, "standalone")))),
  unlist(lapply(runs,           function(r) dc_pdfs(dc_root(r, "pipeline"))))
))
FIGS <- FIGS[file.exists(FIGS)]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA
```

**Runs:** pilot `` `r pilot` `` (where the parameter sweep was done), then
`` `r paste(short(runs), collapse = ", ") `` through the pipeline.

**Status:** settled. `positive.quantile = 0.99` was adopted and is what the Nextflow pipeline
runs for every sample.

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `07.demux_benchmark_report.Rmd`, from the outputs of
steps 01-06 in this folder.

Samples are pooled several to a GEM well, so every cell has to be assigned back to a donor.
Two independent methods are available: hashtag oligos (HTO), which are added at the bench, and
genotype demultiplexing from the RNA itself via cellsnp-lite and vireo. This report compares
them.

For full report, download `demux_benchmark_report.html` in the attachment and open in a
browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/Benchmarking-demultiplexing_with_genotyping_vs_HTO>

# Aim

Decide which demultiplexing method to carry into the pipeline, and on what setting:

- do HTO hashing and genotype demultiplexing recover the same donors
- which assigns more cells as confident singlets, and which calls more doublets
- what `positive.quantile` threshold should HTODemux use
- does the method still work for a donor that was never hashed

# Conclusions

- genotyping with cellSNP/vireo outperforms HTO hashing, and is the method the pipeline now
  relies on for donor assignment
- HTO is still worth running: it is what names the donors, since vireo only produces anonymous
  `donor0…donorN` labels that have to be matched back to sample identity
- `positive.quantile = 0.99` was adopted after the sweep and is what every pipeline run uses;
  the pilot's original 0.80 is too permissive
- the unhashed-donor case works after the fix in `nextflow/bin/process_cellsnp_hto.R`, verified
  against LK3's real output rather than a reimplementation

# 1. How the two methods are produced

HTO comes from CITE-seq-Count on the hashtag library (step 01), whitelisted to the emptyDrops
barcodes, then `HTODemux`. Genotype demultiplexing (step 02) runs cellsnp-lite over the same
barcode list against the Cell Ranger BAM, then vireo with `N_DONORS=4`. Step 03 joins the two
and produces everything below.

The important asymmetry: **vireo output is anonymous.** It returns `donor0`, `donor1` … with no
idea which is which. HTO is what supplies the mapping from genotype cluster to sample name,
which is why both are run even though genotyping is the better classifier. That mapping is
saved as `vireo_to_hto_sample_name_map_positive_quantile_0.99.csv`.

```{r gt-distance, results='asis'}
for (r in c(pilot, runs)) {
  p <- file.path(res, "genotype_demux", r)
  if (!dir.exists(p)) next
  f <- list.files(p, pattern = "fig_GT_distance_estimated\\.pdf$", recursive = TRUE, full.names = TRUE)
  if (length(f)) show_fig(f[1], paste0("Vireo estimated genotype distance between donors, ",
    short(r), ". Confirms the donors are genetically distinguishable before any comparison is made."),
    width_pct = 70)
}
```

# 2. The parameter sweep

The sweep was done on the pilot run. `HTODemux`'s `positive.quantile` sets how aggressively a
cell is called positive for a hashtag; every comparison figure was produced at the default
0.80, at 0.99, and across all quantiles.

```{r sweep-figs, results='asis'}
d <- dc_root(pilot, "standalone")
if (!is.na(d)) {
  for (f in c("01_demux_efficiency_barplot", "03_hto_vireo_donor_heatmap",
              "04_upset_HTO_vireo_overlap")) {
    for (suffix in c("", "_all_positive_quantiles")) {
      p <- dc_file(d, paste0(f, suffix, ".pdf"))
      lab <- if (suffix == "") "at the default 0.80" else "across all positive quantiles"
      if (file.exists(p)) show_fig(p, paste0("`", f, "` ", lab, "."))
    }
  }
}
```

```{r sweep-tables, results='asis'}
if (!is.na(d)) {
  show_table(read_if(dc_file(d, "demux_efficiency_summary.csv")),
             caption = "Demux efficiency, pilot run, default quantile")
  show_table(read_if(dc_file(d, "demux_efficiency_summary_all_positive_quantiles.csv")),
             caption = "Demux efficiency across the quantile sweep", n = 30)
}
```

Only the 0.99 outputs survive for the later runs, which is the practical record of the
decision: from LK2 onwards nothing else was generated.

# 3. Agreement, per run

```{r agreement, results='asis'}
for (r in runs) {
  d2 <- dc_root(r, "pipeline"); d1 <- dc_root(r, "standalone")
  d <- if (!is.na(d2)) d2 else d1
  if (is.na(d)) next
  cat("\n## ", short(r), "\n\n", sep = "")
  cat(if (identical(d, d2)) "From the Nextflow pipeline.\n\n" else "Standalone run.\n\n")
  show_table(read_if(dc_file(d, "demux_efficiency_summary.csv")),
             caption = paste("Demux efficiency -", short(r)))
  show_table(read_if(dc_file(d, "hto_vireo_agreement_summary.csv")),
             caption = paste("HTO / vireo agreement -", short(r)))
  show_table(read_if(dc_file(d, "vireo_summary.csv")),
             caption = paste("Vireo summary -", short(r)))
}
```

# 4. The settled comparison

Figures at `positive.quantile = 0.99`, as the pipeline produces them.

```{r settled-figs, results='asis'}
for (r in runs) {
  d <- dc_root(r, "pipeline")
  if (is.na(d)) next
  cat("\n## ", short(r), "\n\n", sep = "")
  notes <- c(
    "01_demux_efficiency_barplot" = "Singlet / doublet / negative rates for HTO and vireo side by side. The headline comparison.",
    "02_hto_vireo_doublet_comparison" = "Whether the two methods call the same cells doublets.",
    "03_hto_vireo_donor_heatmap" = "HTO sample against vireo donor. Off-diagonal mass is the disagreement; this is also the figure the sample-name mapping is read off.",
    "04_upset_HTO_vireo_overlap" = "UpSet of the call overlap.",
    "05_HTO_QC" = "HTO QC at the adopted threshold.",
    "06_HTO_ridge" = "Ridge plots per hashtag.",
    "06b_HTO_ratio_thresholds" = "HTO ratio thresholds.",
    "07_UMAP_HTO_cellSNP" = "Both sets of calls on the UMAP."
  )
  for (nm in names(notes)) {
    p <- dc_file(d, paste0(nm, "_positive_quantile_0.99.pdf"))
    if (!is.na(p)) show_fig(p, notes[[nm]])
  }
}
```

# 5. The unhashed donor

Not every donor in a pool carries a hashtag. Step 06 checks that case: it runs the production
script `nextflow/bin/process_cellsnp_hto.R` with exactly the arguments the pipeline module
would pass for LK3-GEX, against that run's already-completed cellSNP/vireo and HTO output.
Running the real script rather than a reimplementation is the point - it exercises the code
that ships.

The trial output lives in `scripts/TRIAL_output_unhashed_donor_demux/`, deliberately outside
`nextflow/`, following the convention that trial and diagnostic scripts never sit inside the
pipeline directory.

# 6. What was run, in order

| Step | Script | What it does |
|---|---|---|
| 01 | `01.hto_count.sbatch` | CITE-seq-Count on the HTO library, whitelisted to emptyDrops barcodes. |
| 02 | `02.cellSNP_vireo.sh` | cellsnp-lite pileup over the same barcodes, then vireo with `N_DONORS=4`. |
| 03 | `03.process_cellSNP_HTO.R` | Joins vireo onto the object, cross-tabulates against HTO, writes the sweep. |
| 04 | `04.QC_HTO_ADT_only.R` | QC restricted to the HTO and ADT modalities. |
| 05 | `05.QC_HTO_comparison.R` | Compares HTO and ADT behaviour between the LK1 and LK2 runs. |
| 06 | `06.test_unhashed_donor_demux.sh` | Verifies the unhashed-donor fix against LK3's real output. |
| 07 | `07.demux_benchmark_report.Rmd` | This report. |

The biotin-specific part of step 05 was later split into `scripts/6b1.biotin_ADT_comparison.R`,
which belongs to the biotin report rather than here.

# 7. Data locations

| What | Location |
|---|---|
| Genotype demux | `results/genotype_demux/<run>/<sample>/cellsnp/` and `.../vireo/` |
| Comparison, standalone | `results/demux_comparison/<run>/<sample>/` |
| Comparison, pipeline | `results_nf/<run>/06_demux_comparison/` |
| HTO counts | `results/cite_seq_count/<run>/hto_counts_*/` |
| Unhashed-donor trial | `scripts/TRIAL_output_unhashed_donor_demux/` |

```{r locations, results='asis'}
rows <- list()
for (r in c(pilot, runs)) {
  for (lab in c("standalone", "pipeline")) {
    d <- dc_root(r, lab)
    if (is.na(d)) next
    f <- list.files(d, recursive = TRUE, full.names = TRUE)
    f <- f[!grepl("(^|/)\\._", f)]
    if (!length(f)) next
    dt <- as.Date(file.info(f)$mtime)
    rows[[length(rows) + 1]] <- tibble(run = short(r), source = lab,
                                       files = length(f), first = min(dt), last = max(dt))
  }
}
show_table(bind_rows(rows), caption = "Demux comparison outputs by run and source")
```

# 8. Tools

| Tool | Version | Source |
|---|---|---|
| cellsnp-lite | 1.2.3 (htslib 1.23.1) | `cellsnp_env` |
| vireo (vireoSNP) | to confirm | `cellsnp_env` |
| CITE-seq-Count | to confirm | `01.hto_count.sbatch` |
| Seurat `HTODemux` | see session below | `positive.quantile = 0.99` adopted |

```{r session}
sessionInfo()
```

# 9. Caveats

1. **The sweep exists only for the pilot run.** From LK2 onwards only the 0.99 outputs were
   generated, so the comparison across thresholds cannot be redone from saved files for the
   later runs without re-running step 03.
2. **`results/demux_comparison/` has no entry for LK1-resequenced or LK3.** Those runs were
   demultiplexed through the pipeline, so their comparison lives under `results_nf/<run>/`.
3. Vireo's donor labels are anonymous; any statement tying a genotype cluster to a sample name
   depends on the HTO mapping in `vireo_to_hto_sample_name_map_positive_quantile_0.99.csv`.
