---
title: "Biotin report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  split_sample: "WEI21_26-17_NK"
---

<!--
# ------------------------------------------------------------------
# Biotin report - step 04 of 04
#
# Renders the lab archive report from the outputs of steps 01-03.
# Render with ./render_report.sh, which also prints the HTML to PDF.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))
options(scipen = 999)

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) {
    cat(sprintf("\n_(could not rasterise `%s`: %s)_\n\n",
                basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

# slide-derived PNGs sit next to this file
show_png <- function(rel, note = NULL, width_pct = 100) {
  if (!file.exists(rel)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(rel)))
  if (!is.null(note)) cat(note, "\n\n")
  cat(sprintf('<img src="%s" style="width:%d%%"/>\n\n', rel, width_pct))
}

read_if <- function(path, ...) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj <- params$proj
res  <- file.path(proj, "results")
split_dir <- file.path(res, "hsc_mpp_biotin_split")
mono_dir  <- file.path(res, "monocyte_biotin_split")
cross_dir <- file.path(res, "seurat_demux_test")

find_one <- function(root, name) {
  if (!dir.exists(root)) return(NA_character_)
  hits <- list.files(root, pattern = paste0("^", gsub("\\.", "\\\\.", name), "$"),
                     recursive = TRUE, full.names = TRUE)
  hits <- hits[!grepl("(^|/)\\._", hits)]
  if (!length(hits)) NA_character_ else hits[1]
}

FIGS <- c(list.files(split_dir, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE),
          list.files(mono_dir,  pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE),
          list.files(cross_dir, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE))
FIGS <- FIGS[!grepl("(^|/)\\._", FIGS) & file.exists(FIGS)]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA
```

**Samples:** biotin was compared across LK1 and LK2, and the high/low split was done on
`` `r params$split_sample` `` from LK3.

**Status:** the labelling works. The high-versus-low split is exploratory.

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `04.biotin_report.Rmd`, from the outputs of steps 01-03
in this folder.

Biotin is carried on the ADT panel as a readout of cell-surface glycosylation - "sugar-seq".
It was added deliberately from the pilot run onward, so every CITE-seq experiment carries it
alongside the antibody panel. This report covers how much biotin signal there is, how it is
distributed, and what separates biotin-high from biotin-low cells.

For full report, download `biotin_report.html` in the attachment and open in a browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/6.Biotin_report>

# Aim

- does the biotin labelling work at all, and is it consistent between runs
- how is biotin distributed across cell types
- what distinguishes biotin-high from biotin-low cells within one population

# Conclusions

- **sugar-seq worked** - biotin signal is present and structured in every sample, not uniform
  noise
- **biotin is strongly cell-type dependent.** Monocytes carry markedly the most, with
  pro-monocytes and cDC next; lymphoid populations and erythroid precursors carry the least.
  The pattern holds in both normal donors and the AML sample
- the signal is spatially coherent on the projected UMAP rather than scattered, which is what
  rules out its being a technical artefact of the ADT panel
- within HSC/MPP, splitting on biotin gives differences in RNA, in surface protein and in
  cell-cycle phase, but this arm is exploratory - it has been run on one sample only
- the same split in monocytes runs but on only 160 cells, and does not separate cycling from
  non-cycling: both groups are ~90% G1
- glycosylation has to be read against total surface load: step 02 computes CITE residuals
  adjusted for that, because otherwise a cell with more surface protein of any kind looks
  more glycosylated

# 1. Does it work, and is it consistent

Step 01 compares biotin ADT levels between the LK1 and LK2 runs. It was split out of the
broader HTO/ADT comparison so it can be run without the demultiplexing steps.

```{r cross-run, results='asis'}
for (f in list.files(cross_dir, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE)) {
  if (grepl("(^|/)\\._", f)) next
  show_fig(f, "Biotin ADT compared across the LK1 and LK2 runs.")
}
```

```{r sugar-seq, results='asis'}
show_png("figures_from_slides/biotin_umap_per_sample.png",
  paste("Biotin on the projected UMAP for the two normal donors and the AML sample. The",
        "signal concentrates in the same region of the map in all three, which is the",
        "evidence that the labelling worked - a technical artefact would not localise",
        "consistently to one part of the manifold across independent samples."))
```

# 2. Distribution across cell types

```{r by-celltype, results='asis'}
show_png("figures_from_slides/biotin_by_celltype_violin.png",
  paste("Biotin by broad cell type, split by sample. Monocytes are the clear outlier, with",
        "pro-monocytes and cDC behind them; Pro-B, Pre-B and late erythroid sit lowest. The",
        "ordering is the same in normal-02, normal-CD34-01 and WEI21_26-17_NK, so it tracks",
        "cell identity rather than sample or disease state."))
show_fig(find_one(split_dir, "3c_glyco_by_fine_celltype_violin.pdf"),
  "The same picture at the fine cell type level, from step 02.")
```

# 3. Biotin-high versus biotin-low in HSC/MPP

Step 02 takes the HSC/MPP compartment, splits it on CITE_DSB biotin, and asks what else
differs. Everything in this section is one sample, `` `r params$split_sample` ``.

```{r split-umaps, results='asis'}
show_fig(find_one(split_dir, "5a_biotin_group_umap_projected.pdf"),
  "The two groups on the projected UMAP.")
show_fig(find_one(split_dir, "5b_biotin_value_umap_projected.pdf"),
  "Biotin as a continuous value on the same layout - the split is a threshold on this.")
```

## RNA

```{r split-rna, results='asis'}
show_fig(find_one(split_dir, "2_pseudobulk_deseq2_volcano_biotin_high_vs_low.pdf"),
  "Pseudobulk DESeq2, biotin-high against biotin-low.")
show_fig(find_one(split_dir, "2_pseudobulk_edgeR_volcano_biotin_high_vs_low.pdf"),
  "The same contrast under edgeR - shown alongside because with one sample the choice of method matters more than usual.")
show_fig(find_one(split_dir, "1b_fgsea_hallmark_biotin_high_vs_low.pdf"),
  "fgsea over MSigDB Hallmark, on the FindMarkers ranking.")
show_fig(find_one(split_dir, "5d_findmarkers_top_genes_umap_projected.pdf"),
  "Top FindMarkers genes on the projected UMAP.")
show_fig(find_one(split_dir, "6a_findmarkers_top_genes_violin.pdf"),
  "The same genes as violins by group.")
```

## Surface protein

```{r split-protein, results='asis'}
show_fig(find_one(split_dir, "5f_cite_dsb_top_proteins_umap_projected.pdf"),
  "Top differential CITE_DSB proteins on the projected UMAP.")
show_fig(find_one(split_dir, "6c_cite_dsb_top_proteins_violin.pdf"),
  "The same proteins as violins by group.")
show_fig(find_one(split_dir, "3b_glyco_residual_vs_protein_summary.pdf"),
  paste("Glycosylation against total surface protein, as residuals. This is the control that",
        "matters: without adjusting for surface load, any cell carrying more protein of any",
        "kind would appear more glycosylated."))
show_fig(find_one(split_dir, "3b_glyco_residual_vs_protein_top_scatter.pdf"),
  "The individual proteins behind that summary.")
```

## Cell cycle

```{r split-cc, results='asis'}
show_fig(find_one(split_dir, "4_cell_cycle_phase_proportions_biotin_high_vs_low.pdf"),
  "Cell-cycle phase proportions in the two groups, from Seurat CellCycleScoring.")
show_fig(find_one(split_dir, "5c_cell_cycle_phase_umap_projected.pdf"),
  "Phase on the projected UMAP, to check the split is not simply tracking proliferation.")
```

```{r split-tables, results='asis'}
for (f in list.files(split_dir, pattern = "\\.csv$", full.names = TRUE)) {
  if (grepl("(^|/)\\._", f)) next
  show_table(read_if(f), caption = basename(f), n = 15)
}
```

# 3b. The same split in monocytes

Monocytes carry the most biotin, so they are the obvious population to split. This arm had
never been run - the report script existed but nothing wrote the object it needs. Step 02a is
step 02 with the population taken as an argument; run as
`Rscript 02a.biotin_split_analysis_runnable.R "Monocyte"` it produced everything below.

**The population is small.** 160 monocytes pass QC in this sample, splitting 80 high and 80
low at the median. That is a tenth of the HSC/MPP compartment, and it is why the pseudobulk
contrast here is thinner - treat the gene-level results as indicative only.

```{r mono-figs, results='asis'}
show_fig(find_one(mono_dir, "5a_biotin_group_umap_projected.pdf"),
  "Biotin-high and biotin-low monocytes on the projected UMAP.")
show_fig(find_one(mono_dir, "5b_biotin_value_umap_projected.pdf"),
  "Biotin as a continuous value across the same cells.")
show_fig(find_one(mono_dir, "2_pseudobulk_deseq2_volcano_biotin_high_vs_low.pdf"),
  "Pseudobulk DESeq2, biotin-high against biotin-low monocytes.")
show_fig(find_one(mono_dir, "1b_fgsea_hallmark_kegg_biotin_high_vs_low.pdf"),
  "fgsea over Hallmark and KEGG on the FindMarkers ranking.")
show_fig(find_one(mono_dir, "3b_glyco_by_fine_celltype_violin.pdf"),
  "Glycosylation by fine cell type within the monocyte compartment.")
show_fig(find_one(mono_dir, "3c_glyco_residual_vs_protein_summary.pdf"),
  "Glycosylation against total surface load, as residuals.")
show_fig(find_one(mono_dir, "4_cell_cycle_phase_proportions_biotin_high_vs_low.pdf"),
  paste("Cell-cycle phase by group. Both are overwhelmingly G1 - 92.5% of biotin-high and",
        "87.5% of biotin-low - so the split is not separating cycling from non-cycling cells."))
show_fig(find_one(mono_dir, "6c_cite_dsb_top_proteins_violin.pdf"),
  "Top differential CITE_DSB proteins between the groups.")
```

```{r mono-tables, results='asis'}
for (nm in c("3b_fine_celltype_counts.csv",
             "4_cell_cycle_phase_proportions_biotin_high_vs_low.csv",
             "3_cite_dsb_wilcoxon_biotin_high_vs_low.csv")) {
  show_table(read_if(find_one(mono_dir, nm)), caption = nm, n = 15)
}
```

Two differences from the HSC/MPP outputs are worth noting rather than glossing. This run
writes its UMAPs into a `umaps/` subdirectory while the HSC/MPP figures sit flat, and it
produces no edgeR volcano where HSC/MPP has one - both because the HSC/MPP outputs came from
an earlier version of the analysis script than the one frozen here as step 02.

# 4. What was run, in order

| Step | Script | What it does |
|---|---|---|
| 01 | `01.biotin_ADT_comparison.R` | Biotin ADT compared across the LK1 and LK2 runs. |
| 02a | `02a.biotin_split_analysis_runnable.R` | Step 02 with the population as an argument, so any cell type can be split. Used to produce the monocyte arm. |
| 02 | `02.hsc_mpp_biotin_split_analysis.R` | HSC/MPP split on biotin, then RNA DE, fgsea, pseudobulk, CITE DE, glycosylation by cell type, surface-load residuals, cell cycle. |
| 03 | `03.biotin_split_report.Rmd` | The report over step 02's outputs. |
| 04 | `04.biotin_report.Rmd` | This report. |

Three scripts are not kept as separate steps. `scripts/27b.hsc_mpp_biotin_split_report.Rmd` is
the longer first version of step 03 and is superseded by it.
`scripts/27d.monocyte_biotin_split_report_short.Rmd` is step 03 pointed at monocytes,
differing by three lines - the title and two paths. It could not run before because nothing
wrote the object it reads; step 02a now does, so it is reproducible. `scripts/27.removeme_quick_biotin_test.R` is scratch, named by its author as
disposable.

# 5. Data locations

| What | Location |
|---|---|
| Cross-run biotin comparison | `results/seurat_demux_test/LK1_vs_LK2_test_HTO_ADT_comparison/` |
| HSC/MPP split | `results/hsc_mpp_biotin_split/` |
| Monocyte split | `results/monocyte_biotin_split/`, figures under `umaps/` |
| Rendered prior reports | `scripts/27b.hsc_mpp_biotin_split_report.html`, `scripts/27c.hsc_mpp_biotin_split_report_short.html` |
| Slide figures | `figures_from_slides/`, from the 2026-07-22 progress deck |

# 6. Tools

| Tool | Version | Source |
|---|---|---|
| Seurat `FindMarkers` | see session below | Wilcoxon, single cell |
| DESeq2 / edgeR | see session below | pseudobulk, fixed dispersion |
| fgsea + msigdbr | see session below | Hallmark gene sets |
| Seurat `CellCycleScoring` | see session below | phase assignment |

```{r session}
sessionInfo()
```

# 7. Caveats

1. **The high/low split is one sample.** Everything in section 3 comes from
   `` `r params$split_sample` `` alone, so the RNA and protein differences are not replicated.
   Treat them as hypotheses.
2. **The monocyte arm rests on 160 cells**, 80 high and 80 low. It has now been run, but that
   is a tenth of the HSC/MPP compartment, so its gene-level results are weaker still.
3. **Both DESeq2 and edgeR volcanoes are shown** rather than one, because with a single sample
   the pseudobulk dispersion is fixed rather than estimated and the two methods can disagree.
4. **The cross-run comparison is LK1 against LK2 only.** LK3, which supplied the split sample,
   is not in it.
