---
title: "Nextflow workflow"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
---

<!--
# ------------------------------------------------------------------
# Nextflow workflow - step 03 of 03
#
# Renders the lab archive report from the outputs of steps 01-02.
# Render with ./render_report.sh.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))
options(scipen = 999)

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) {
    cat(sprintf("\n_(could not rasterise `%s`: %s)_\n\n",
                basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

read_if <- function(path, ...) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

# Pandoc reads $...$ as inline math, so a published path like
# ${run}/08_annotate turns the underscore into a subscript. Wrapping such cells
# in backticks makes them inline code, which pandoc leaves alone. Applied to any
# column whose values contain $ or _, plus anything named in `code`.
as_code <- function(x) {
  x <- as.character(x)
  ifelse(is.na(x) | x == "", x, paste0("`", x, "`"))
}

needs_code <- function(x) {
  x <- as.character(x)
  any(grepl("[$_]", x[!is.na(x)]))
}

show_table <- function(df, caption = NULL, n = Inf, code = NULL) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  df <- as.data.frame(df)
  for (nm in names(df)) {
    if (nm %in% code || (is.null(code) && needs_code(df[[nm]]))) {
      df[[nm]] <- as_code(df[[nm]])
    }
  }
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj      <- params$proj
bench_dir <- file.path(proj, "results", "benchmarking", "nextflow_workflow")

all_pdfs <- function(root) {
  if (!dir.exists(root)) return(character(0))
  f <- list.files(root, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE)
  f[!grepl("(^|/)\\._", f)]
}

FIGS <- all_pdfs(bench_dir)
FIGS <- FIGS[file.exists(FIGS) & file.info(FIGS)$size > 0]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA

t_proc   <- read_if(file.path(bench_dir, "01_pipeline_processes.csv"))
t_order  <- read_if(file.path(bench_dir, "02_workflow_order.csv"))
t_params <- read_if(file.path(bench_dir, "03_pipeline_params.csv"))
t_res    <- read_if(file.path(bench_dir, "04_process_resources.csv"))
t_sheet  <- read_if(file.path(bench_dir, "05_samplesheet.csv"))
t_obs    <- read_if(file.path(bench_dir, "06_pipeline_observations.csv"))
t_wired  <- read_if(file.path(bench_dir, "07_process_wired_into_workflow.csv"))
t_stages <- read_if(file.path(bench_dir, "09_stage_presence_by_run.csv"))
t_funnel <- read_if(file.path(bench_dir, "10_cell_count_funnel.csv"))
t_lost   <- read_if(file.path(bench_dir, "11_cells_lost_by_step.csv"))
t_ovl    <- read_if(file.path(bench_dir, "12_emptydrops_cellbender_overlap.csv"))
t_cmp    <- read_if(file.path(bench_dir, "13_params_vs_benchmarks.csv"))
t_runmap <- read_if(file.path(bench_dir, "14_run_identity.csv"))
t_rtproc <- read_if(file.path(bench_dir, "16_runtime_by_process_by_run.csv"))
t_rtwide <- read_if(file.path(bench_dir, "17_runtime_by_process_wide.csv"))
t_rtrun  <- read_if(file.path(bench_dir, "18_runtime_by_run.csv"))
t_elapsed<- read_if(file.path(bench_dir, "20_elapsed_vs_work.csv"))
```

**Runs processed:** `260522_VH01624_461_222JLJVNX` (LK1 re-sequenced),
`260528_VH01624_464_222K7VKNX` (LK2) and `260717_VH01624_477_222KG22NX` (LK3).

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `03.nextflow_workflow_report.Rmd`, from the outputs
of steps 01-02 in this folder.

Everything the benchmark reports settled - cell calling, demultiplexing, ADT normalisation,
annotation, CNV calling - is implemented once, as a Nextflow pipeline, and run over every
sequencing run from raw FASTQ to a final annotated object. This report describes what the
pipeline is, what it produced, and where its settings agree or disagree with the benchmarks
that were supposed to determine them.

For full report, download `nextflow_workflow_report.html` in the attachment and open in a
browser.

The pipeline lives in a **different repository** from the lab archives:
<https://github.com/nbartonicek/surface_proteome>

Code used for this report - attached and in:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/9.Nextflow_workflow>

# Aim

Document the production pipeline as it actually stands, and check it against the benchmarks:

- what does the workflow do, in what order, with what resources
- does every process actually run
- do the parameters match what the benchmark reports concluded
- what did it produce for each run, and where do cells get lost

# Conclusions

- **the pipeline runs end to end on all three runs.** Every one of the 13 published stages
  is present for LK1 re-sequenced, LK2 and LK3, through to a final merged object and a
  per-run QC report
- **most settings match their benchmarks** - emptyDrops FDR 0.01 and minimum 100 UMI, HTO
  `positive.quantile` 0.99, cellSNP min count 20 and MAF 0.10, genotype demultiplexing with
  cellSNP-lite plus vireo, DSB for ADT, CellBender 0.3.0, `refdata-gex-GRCh38-2024-A`
- **the annotation threshold does not match.** `bin/annotate.R` runs BoneMarrowMap at
  `MAD_threshold = 4`; `bin/lib_early_qc_annotation.R` and every standalone script use the
  default 2.5. Report 07 shows this roughly halves the rejection rate, so `results_nf/` and
  `results/seurat_annotated/` are **not** directly comparable
- **`params.cell_caller` is a dead parameter that says the wrong thing.** It is set to
  `"emptydrops"` and is never read anywhere. The workflow assigns CellBender barcodes as the
  primary cell calls unconditionally. Anyone changing that parameter expecting an effect
  would get none
- **`DOROTHEA_VIPER` never runs.** The module and its R script exist, but the process is the
  only one of the 20 that is neither included nor called in `main.nf`
- **demultiplexing is where the cells go.** It is the largest single loss in every run - LK2
  drops from 33,228 scDblFinder singlets to 25,690 demultiplexed cells, about 23%. Cell
  calling and mapping QC are minor by comparison
- **the two cell callers disagree more than expected, and not evenly.** Jaccard overlap is
  91.9% for LK1 but only **74.5%** for LK2, and CellBender always calls more barcodes -
  8,895 CellBender-only against 1,349 emptyDrops-only in LK2. Since CellBender is the primary
  caller and emptyDrops feeds the DSB background, the two are both in use and their
  disagreement propagates
- **the pipeline runs the re-sequenced LK1, not the pilot.** `260522` is in the samplesheet;
  `260423` is not. Reports 01, 07 and 08 analysed the pilot. The two trees look
  interchangeable and are not
- **donor names differ between the pipeline and the standalone work** - `HBDN206-MN-pCT`
  against `HBDN206-MNpCT`, `HBDN392-MDS` against `HBDN392-AML-MDS`, `HBDN501-KMT2A` against
  `HBDN501-AML-KMT2A`. Joins between the two output trees will not match on donor

# 1. What the pipeline is

A single `main.nf` with `r if (!is.null(t_proc)) nrow(t_proc) else "18"` modules, run on SLURM
through `run_nextflow.sh`, publishing to `results_nf/<run>/`. Each process is a thin Nextflow
wrapper around an R script in `bin/`, so the analysis code is readable on its own and the
`.nf` file carries only the plumbing.

```{r sheet, results='asis'}
show_table(t_sheet, "conf/samples.csv - the only thing that has to change to add a run")
```

The samplesheet carries the run, the sample, the donor count and the semicolon-joined donor
IDs. LK3's `donor_ids` field - `normal_CD34-01;;;normal-02` - shows two conventions at once:
the empty slots are HTO positions that were not used, and `WEI21_26-17_NK` is an *unhashed*
donor carried in its own column, assigned downstream by whichever vireo genetic cluster does
not match an HTO singlet population.

```{r order, results='asis'}
show_table(t_order, "Workflow order, from the section comments in main.nf")
```

```{r procs, results='asis'}
if (!is.null(t_proc)) {
  show_table(t_proc %>% dplyr::select(module, process, publishes),
             "Modules, the processes they define, and where they publish")
}
```

# 2. Does every process run

```{r wired, results='asis'}
if (!is.null(t_wired)) {
  show_table(t_wired %>% arrange(included_in_main, process),
             "Every process declared in modules/, and whether main.nf includes it")
}
```

`DOROTHEA_VIPER` is declared in `modules/dorothea_viper.nf` with a matching
`bin/dorothea_viper.R`, and is neither included nor invoked. It is dead code rather than a
disabled option - there is no parameter that would switch it on.

# 3. Parameters against the benchmarks

This is the part that matters. Each benchmark report exists to settle one setting; the
pipeline is where those decisions are supposed to end up.

```{r cmp, results='asis'}
show_table(t_cmp, "Pipeline setting against the benchmark that determined it", code = "pipeline_value")
```

Two rows need attention.

**`MAD_threshold`.** `bin/annotate.R` - the script the live `ANNOTATE` process runs - uses 4.
`bin/lib_early_qc_annotation.R` uses 2.5, and so did every standalone annotation script.
Report 07 measured what this costs: rejection drops from 7.95% to 3.91% in the pilot and from
5.48% to 1.75% in LK2. The cells recovered are shallow, ribosome-poor and enriched for
HTO negatives, which argues for 2.5. Whichever is preferred, the two code paths should agree
with each other, and right now they do not.

**`params.cell_caller`.** Declared as `"emptydrops"` in `nextflow.config` and never read. The
workflow does `cell_barcodes = CELLBENDER.out.cellbender_barcodes` with no conditional. The
behaviour is right - report 04 concluded CellBender for cell calls with emptyDrops retained
for the DSB background, which is exactly what the DAG does - but the parameter documents the
opposite and would mislead anyone reading the config to find out what the pipeline does.

```{r params, results='asis'}
show_table(t_params, "All pipeline parameters as configured")
```

# 4. Resources

```{r res, results='asis'}
show_table(t_res, "Per-process resource requests")
```

Defaults are 2 CPUs, 8 GB and 2 hours; `errorStrategy` is `finish`, so a failing process lets
already-running work complete rather than killing the run, but does not retry.

# 5. What it produced

```{r stages, results='asis'}
show_table(t_stages, "Published stage directories, entries per run")
```

All three runs completed all 13 stages.

```{r funnel, results='asis'}
show_table(t_funnel %>% dplyr::select(run, sample, step, n_cells),
           "Cells surviving each stage")
show_fig(file.path(bench_dir, "01_cell_count_funnel.pdf"),
  paste("EmptyDrops and CellBender sit at the same point in the funnel - they are",
        "alternatives, not sequential filters, which is why the line rises between them."))
```

```{r lost, results='asis'}
show_table(t_lost %>% dplyr::select(run, sample, step, n_cells_lost),
           "Cells lost at each transition")
```

Demultiplexing is consistently the largest loss. In LK2 it removes 7,538 cells, against 348
lost to mapping QC at the other end. That is the expected shape - pooled donors mean every
ambiguous or doublet droplet is discarded - but it is worth knowing that the dominant
attrition in the pipeline is the pooling design, not any QC threshold.

# 6. The two cell callers

```{r overlap, results='asis'}
show_table(t_ovl %>% dplyr::select(run, sample, n_emptydrops, n_cellbender, n_common,
                                   n_emptydrops_only, n_cellbender_only, pct_jaccard),
           "Barcode overlap between the two callers")
show_fig(file.path(bench_dir, "02_cell_caller_overlap.pdf"))
```

Both callers run on every sample, and both outputs are used - CellBender's barcodes are the
cell calls, emptyDrops' barcodes define the background droplets DSB needs. So the overlap is
not academic: the two sets disagree by 25% in LK2, and each disagreement region feeds a
different part of the downstream analysis.

# 7. Runtime

Nextflow writes no trace file here - only `report` and `dag` are enabled in the config - but
its HTML execution report embeds the full per-task trace as JSON, which is where these
numbers come from. They are Nextflow's own accounting, not a re-measurement.

```{r rt-run, results='asis'}
show_table(t_rtrun, "Per run. wall_total_if_serial is the sum of task runtimes, not elapsed time")
show_table(t_elapsed, "Elapsed wall clock against work done. concurrency = work / elapsed")
```

`CACHED` is a success, not a failure - it means Nextflow resumed that task from a previous
run's work directory instead of re-executing it. LK1 was executed in full; LK2 and LK3 are
almost entirely resumed. **No task failed in any run.**

```{r timeline, results='asis'}
show_fig(file.path(bench_dir, "06_pipeline_timeline.pdf"),
  paste("One bar per task, positioned from the first submit of its run. Overlapping bars ran",
        "concurrently, so panel width is elapsed time while the total bar length is work",
        "done. The long serial spine at the start - Cell Ranger, then CellBender and",
        "cellSNP - is what sets the floor on how fast a run can finish."), dpi = 110)
```

```{r rt-proc, results='asis'}
if (!is.null(t_rtproc)) {
  show_table(
    t_rtproc %>%
      dplyr::select(run, process, n_tasks, cached, wall_total, wall_longest,
                    cpu_hours, peak_rss_gb),
    "Per process per run, longest first"
  )
}
show_table(t_rtwide, "Wall-clock time per process, runs side by side")
```

```{r rt-figs, results='asis'}
show_fig(file.path(bench_dir, "03_runtime_by_process.pdf"))
show_fig(file.path(bench_dir, "04_cpu_hours_by_process.pdf"),
  "Allocated CPU hours - realtime times allocated cpus, so what the process costs the cluster rather than what it uses.")
show_fig(file.path(bench_dir, "05_peak_memory_by_process.pdf"),
  paste("Peak RSS per task against the memory each process asks for. CITE_SEQ_COUNT_ADT is",
        "the one to look at."))
```

Three things stand out. **CopyKAT is the most expensive step** - 10h19m of wall clock across
its four per-donor tasks in LK2, more than Cell Ranger. **The pipeline is only 1.5-2.2x
concurrent**, because the early stages are a serial chain on one sample before the per-donor
fan-out begins. And **CITE_SEQ_COUNT_ADT peaks at 311 GB RSS**, which is worth checking
against what its process directive actually requests.

# 8. Which run is which

```{r runmap, results='asis'}
show_table(t_runmap, "Run identity", code = "run")
```

# 9. Observations

Recorded rather than fixed - the pipeline is a separate repository and nothing in this report
modifies it.

```{r obs, results='asis'}
show_table(t_obs, "Things worth deciding about", code = "where")
```

# 10. What was run

```{r steps, results='asis'}
steps <- tibble::tribble(
  ~Step, ~Script, ~What,
  "01", "01.pipeline_inventory.R",
  "Parses modules/, main.nf and nextflow.config into process, parameter and resource tables, and checks which processes are actually wired into the workflow.",
  "02a", "02a.pipeline_runtimes.R",
  "Parses the per-task trace out of Nextflow's HTML execution reports and summarises runtime, CPU hours, peak memory and the schedule.",
  "02", "02.pipeline_outputs_and_params.R",
  "Inventories the published output tree per run, reads the cell-count funnel and cell-caller overlap, and compares every parameter against the benchmark that determined it.",
  "03", "03.nextflow_workflow_report.Rmd",
  "This report."
)
show_table(steps, code = "Script")
```

Unlike reports 01-08 there are no analysis scripts of mine to collate here - the pipeline
itself is the artefact. These two steps read it and write down what they find; neither writes
anything into `nextflow/`.

# 11. Data locations

```{r locations, results='asis'}
locs <- tibble::tribble(
  ~What, ~Where,
  "Pipeline", "nextflow/ - separate git repository, nbartonicek/surface_proteome",
  "Entry point", "nextflow/main.nf",
  "Process definitions", "nextflow/modules/*.nf",
  "Analysis code", "nextflow/bin/*.R and qc_report.Rmd",
  "Configuration", "nextflow/nextflow.config",
  "Samplesheet", "nextflow/conf/samples.csv",
  "Launcher", "nextflow/run_nextflow.sh",
  "Published output", "results_nf/<run>/",
  "Nextflow's own DAG and execution reports", "results_nf/pipeline_dag.html, results_nf/execution_report_<run>.html",
  "Tables and figures for this report", "results/benchmarking/nextflow_workflow"
)
show_table(locs)
```

# 12. Caveats

- this report describes the pipeline as it stands on the `reporting` branch of the
  `surface_proteome` repository at the time of writing. It is a live repository and can move
  underneath this description
- the parameter-versus-benchmark table is a judgement about intent, not a measurement. It
  records what each benchmark report concluded and what the pipeline does; where they agree,
  that agreement was not re-verified by re-running anything
- the cell-count funnel and cell-caller overlap tables are the pipeline's own QC output, read
  as published. They were not recomputed from the objects
- `results_nf/` and `results/seurat_annotated/` are different analyses of overlapping data,
  at different annotation thresholds, on partly different runs, with different donor naming.
  Numbers should not be carried between them without checking all three
- runtime numbers come from Nextflow's execution reports. For a CACHED task the timestamps
  are those of the run that actually executed it, not of the resumed invocation, so the
  timelines for LK2 and LK3 show the original schedule
- CPU hours are allocated, not consumed - a process given 16 cores that uses one still bills
  16
- nothing in this report modifies the pipeline. The observations in section 8 are for you to
  decide about
