---
title: "Benchmarking: calling cancer cells"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  pilot_run: "260423_VH01624_453_222HWMYNX"
  run: "260528_VH01624_464_222K7VKNX"
---

<!--
# ------------------------------------------------------------------
# Calling cancer cells - step 09 of 09
#
# Renders the lab archive report from the outputs of steps 01-08.
# Render with ./render_report.sh, which also prints the HTML to PDF.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))
options(scipen = 999)

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) {
    cat(sprintf("\n_(could not rasterise `%s`: %s)_\n\n",
                basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

show_img <- function(path, note = NULL, width_pct = 100) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  b64 <- base64enc::dataURI(file = path, mime = "image/jpeg")
  cat(sprintf('<img src="%s" style="width:%d%%"/>\n\n', b64, width_pct))
}

read_if <- function(path, ...) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj  <- params$proj
res   <- file.path(proj, "results")
run   <- params$run
pilot <- params$pilot_run

bench_dir   <- file.path(res, "benchmarking", "cancer_calling")
copykat_dir <- file.path(res, "seurat_annotated", pilot, "copykat_annotated")
numbat_lk1  <- file.path(res, "seurat_annotated", pilot, "numbat")
numbat_lk2  <- file.path(res, "seurat_annotated", run, "numbat")
lr_fig      <- file.path(res, "longread_numbat", "comparison_figures")
lr_tbl      <- file.path(res, "longread_numbat", "comparison_tables")
normal_dir  <- file.path(proj, "results_nf", run, "09c_numbat_run",
                         "numbat_interpretable_cnv_qc")

find_one <- function(root, name) {
  if (is.null(root) || !dir.exists(root)) return(NA_character_)
  hits <- list.files(root, pattern = paste0("^", gsub("\\.", "\\\\.", name), "$"),
                     recursive = TRUE, full.names = TRUE)
  hits <- hits[!grepl("(^|/)\\._", hits)]
  if (!length(hits)) NA_character_ else hits[1]
}

all_pdfs <- function(root, pattern = "\\.pdf$") {
  if (!dir.exists(root)) return(character(0))
  f <- list.files(root, pattern = pattern, recursive = TRUE, full.names = TRUE)
  f[!grepl("(^|/)\\._", f)]
}

FIGS <- unique(c(all_pdfs(bench_dir), all_pdfs(normal_dir),
                 all_pdfs(copykat_dir), all_pdfs(lr_fig)))
FIGS <- FIGS[!is.na(FIGS) & file.exists(FIGS)]
FIGS <- FIGS[file.info(FIGS)$size > 0]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA

t_ck      <- read_if(file.path(bench_dir, "01_copykat_calls_by_donor.csv"))
t_nb      <- read_if(file.path(bench_dir, "02_numbat_compartment_by_donor.csv"))
t_conf    <- read_if(file.path(bench_dir, "03_copykat_vs_numbat_confusion.csv"))
t_agree   <- read_if(file.path(bench_dir, "04_copykat_vs_numbat_agreement.csv"))
t_agree_ct<- read_if(file.path(bench_dir, "05_copykat_vs_numbat_agreement_by_celltype.csv"))
t_lr      <- read_if(file.path(bench_dir, "06_shortread_vs_longread.csv"))

t_norm_cut <- read_if(file.path(normal_dir, "normal_empirical_cutoffs.csv"))
t_norm_fr  <- read_if(file.path(normal_dir, "tumour_like_fraction_by_metric.csv"))
```

**Runs:** pilot `` `r pilot` `` (HBDN206-MNpCT, HBDN392-AML-MDS, HBDN501-AML-KMT2A,
MOLM13) and `` `r run` `` / LK2 (normal-01, HBDN376-TP53, HBDN498-TP53, APOP576-TP53).

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `09.cancer_calling_report.Rmd`, from the outputs of
steps 01-08 in this folder.

Which cells in a patient sample are the leukaemia. There is no ground truth, so the question
is approached by running two independent CNV callers on the same cells and asking where they
agree, running one of them on a healthy donor to measure the false-positive floor, and
running the same caller on short-read and long-read sequencing of the same cells to see how
much the call depends on the data rather than the method.

For full report, download `cancer_calling_report.html` in the attachment and open in a
browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/8.Benchmarking-calling_cancer_cells>

# Aim

Establish how malignant cells are identified, and how far the calls can be trusted:

- do two independent CNV callers agree on the same cells, and where do they not
- what does a caller do on a donor with no cancer - the false positive floor
- does the call depend on the sequencing platform
- which populations are hardest to call

# Conclusions

- **CopyKAT and Numbat agree on the great majority of cells** -
  `r if (!is.null(t_agree)) paste0(paste(sprintf("%s%%", t_agree$percent_agree), collapse = " and "), " on ", paste(t_agree$donor, collapse = " and ")) else "n/a"`
- **but that headline agreement is inflated and should not be quoted on its own.** It is
  carried by HSC MPP, which is 4,002 of the 4,409 shared cells in HBDN206-MNpCT and agrees
  99.8%. Agreement in the lymphoid compartment is far worse - Naive T 25%, B 51%, CD4 Memory
  T 63% - and those are exactly the populations handed to CopyKAT as its confident-normal
  anchor. CopyKAT calls them diploid partly by construction, so the two callers are not
  independent there
- **where they disagree, it is one-directional: Numbat calls malignant, CopyKAT calls
  normal** (152 against 9 cells in HBDN206-MNpCT, 73 against 2 in HBDN501-AML-KMT2A). Numbat
  is the more liberal caller on this data
- **Numbat is specific.** On the healthy donor its CNV evidence sits at a median
  `cnv_z_abs_mean` of 15.3 against 123-315 in the three AML donors, and only **1.0%** of
  normal-01 cells exceed the 99th percentile of the normal distribution against **97-100%**
  of AML cells. Calling cancer where there is none is not the failure mode to worry about
- **the platform changes the answer more than the caller does.** Same cells, same Numbat,
  short read against long read: compartment concordance is
  `r if (!is.null(t_lr)) paste(sprintf("%.1f%%", t_lr$compartment_concordance), collapse = ", ") else "n/a"`,
  but per-cell pCNV correlation is only
  `r if (!is.null(t_lr)) paste(sprintf("%.2f", t_lr$pcnv_correlation), collapse = ", ") else "n/a"`.
  So the two platforms sort cells into the same bin reasonably often while disagreeing almost
  entirely about how much CNV evidence each cell carries
- **that disagreement is also one-directional** - short read calls tumour where long read
  calls normal (339, 566 and 365 cells) and almost never the reverse (0, 99, 2). Long read
  is the more conservative platform, which is expected given its much lower depth per cell
- **Numbat did not complete for two of the four pilot donors.** HBDN392-AML-MDS and MOLM13
  are 100% `not_called`. Any pilot statement about Numbat rests on two donors
- **MOLM13 is never actually called by either method.** It is an AML cell line with no normal
  cells to anchor a baseline, so it is assigned aneuploid directly. It is flagged in its own
  column so it can be excluded from concordance statistics, and it should be

# 1. The two callers

CopyKAT and Numbat both infer copy number from single-cell RNA, but they use different
evidence and need different inputs.

CopyKAT works from expression alone. It needs a set of cells it can trust to be diploid, and
it is given the lymphoid populations - Naive T, CD4 Memory T, CD8 Memory T, NK and B - on the
grounds that they are not part of the myeloid compartment the leukaemia arises from. It is
run **per donor**, not per GEM well: the well is a pool of four donors and CopyKAT estimates
its baseline from the cells it is handed, so pooling would let one donor's aneuploidy define
another donor's normal.

Numbat adds allele evidence. It needs the Cell Ranger BAM, a phased SNP panel and a
population reference panel, so it runs in two cluster stages inside the Numbat container -
pileup and phasing first (step 02), then the CNV and clone inference (step 03).

```{r callers, results='asis'}
show_table(t_ck, "CopyKAT calls per donor, LK1 pilot")
show_fig(file.path(bench_dir, "01_copykat_aneuploid_by_donor.pdf"))
```

```{r numbat-calls, results='asis'}
if (!is.null(t_nb)) {
  show_table(t_nb %>% dplyr::select(-any_of("run")),
             "Numbat compartment per donor")
}
```

Two donors come back entirely `not_called`. HBDN392-AML-MDS and MOLM13 did not produce usable
Numbat output, so every Numbat statement in this report is based on HBDN206-MNpCT and
HBDN501-AML-KMT2A only.

# 2. Do the two callers agree

```{r agree, results='asis'}
show_table(t_agree, "Cell-level agreement, on cells with a call from both")
show_fig(file.path(bench_dir, "02_copykat_vs_numbat_confusion.pdf"),
  paste("The off-diagonal is asymmetric. Numbat calling malignant where CopyKAT calls normal",
        "is much more common than the reverse."))
```

The headline number is high, and on its own it is misleading. Breaking agreement down by
cell type shows where it actually comes from:

```{r agree-ct, results='asis'}
show_table(t_agree_ct, "Agreement by broad cell type (populations with at least 20 cells)")
```

HSC MPP is 4,002 of the 4,409 shared cells and agrees 99.8%, which is what drags the overall
figure to 96%. The lymphoid populations - the ones the two callers actually have room to
disagree about - sit between 25% and 80%.

This is worth being blunt about: **the lymphoid cells are CopyKAT's normal reference.**
CopyKAT was told they are diploid, so its calling them diploid is not independent evidence.
Where Numbat calls some of those same cells malignant, the disagreement cannot be settled by
this comparison - it is a question about whether the normal anchor was correctly chosen, and
that is what step 05's clone inspection is for.

# 3. What the callers do on a donor with no cancer

normal-01 in LK2 is a healthy donor. Anything called tumour there is a false positive, so it
gives the floor that the AML numbers should be read against. Step 06 compares it against the
three AML donors on the same per-cell CNV metrics.

```{r normal, results='asis'}
if (!is.null(t_norm_fr)) {
  show_table(
    t_norm_fr %>%
      filter(metric %in% c("cnv_z_abs_mean", "cnv_z_burden_100")) %>%
      mutate(across(where(is.numeric), ~ round(.x, 3))),
    "Tumour-like fraction against the normal donor's own distribution"
  )
}
show_fig(find_one(normal_dir, "02_fraction_above_normal_cnv_cutoffs.pdf"),
  paste("Fraction of each donor's cells exceeding the normal donor's cutoffs. The three AML",
        "donors sit at the top, normal-01 at the floor."))
show_fig(find_one(normal_dir, "01_interpretable_per_cell_cnv_metrics.pdf"),
  "The per-cell metrics themselves, normal against AML.")
show_fig(find_one(normal_dir, "03_sample_level_evidence.pdf"),
  "Sample-level CNV evidence.")
```

The separation is clean. Median `cnv_z_abs_mean` is 15.3 in normal-01 against 123, 133 and
315 in the AML donors, and 1.0% of normal cells exceed the normal 99th percentile against
97-100% of AML cells. Numbat is not manufacturing cancer out of a healthy marrow.

# 4. Short read against long read

The same LK2 cells were sequenced on both platforms and Numbat was run on each. Same donors,
same barcodes, same caller - so any disagreement is the data, not the method.

```{r lr, results='asis'}
show_table(t_lr, "Short read against long read, per donor")
show_fig(file.path(bench_dir, "03_shortread_vs_longread_tumor_fraction.pdf"),
  "Long read calls fewer cells tumour in every donor.")
show_fig(file.path(bench_dir, "04_shortread_vs_longread_direction.pdf"),
  paste("The disagreement is almost entirely one-directional. Cells that short read calls",
        "tumour and long read calls normal outnumber the reverse by orders of magnitude in",
        "two of the three donors."))
```

Two numbers should be read together here. Compartment concordance looks respectable at 89-94%
for two of the donors, but the per-cell pCNV correlation is 0.12-0.41. The platforms put
cells in the same bin without agreeing on the evidence behind it, which means the compartment
call is robust in a way the per-cell CNV probability is not. Anything downstream that uses
pCNV as a continuous quantity should not be trusted across platforms.

APOP576-TP53 is the worst on both counts - 63.6% concordance and 0.12 correlation. It is also
the sample that loses 30% of its cells to BoneMarrowMap mapping-error QC in the annotation
benchmark. The same sample being the outlier in two unrelated analyses points at the sample
rather than at either method.

```{r lr-per-donor, results='asis'}
lrf <- all_pdfs(lr_fig)
for (f in sort(lrf[grepl("summary_", basename(lrf))])) show_fig(f)
```

# 5. Per-donor detail

```{r per-donor, results='asis'}
for (f in sort(all_pdfs(lr_fig))) {
  if (grepl("summary_", basename(f))) next
  show_fig(f)
}
```

```{r copykat-figs, results='asis'}
ckf <- all_pdfs(copykat_dir)
for (f in sort(ckf)) show_fig(f)
```

# 6. What was run

```{r steps, results='asis'}
steps <- tibble::tribble(
  ~Step, ~Script, ~What,
  "01", "01.copykat_per_donor.R",
  "CopyKAT per donor, lymphoid cells as the confident-normal anchor. MOLM13 assigned rather than called.",
  "02", "02.numbat_pileup_and_phase.sh",
  "Numbat stage 1 on the cluster: pileup and phasing from the Cell Ranger BAM, in the container.",
  "03", "03.numbat_run.sh + 03a.numbat_run.R",
  "Numbat stage 2: CNV and clone inference, one array task per donor.",
  "04", "04.numbat_to_seurat.R",
  "Pulls the per-donor Numbat calls onto the annotated object and plots compartment and clone against the BoneMarrowMap annotation.",
  "05", "05.compare_numbat_copykat.R",
  "Numbat against CopyKAT on one donor, with the clone inspection. Donor and doublet handling are arguments.",
  "06", "06.numbat_normal_donor_control.R",
  "The specificity control: what Numbat reports on the healthy donor, against the three AML donors.",
  "07", "07.compare_shortread_longread_numbat.R",
  "Short-read against long-read Numbat on the same cells.",
  "08", "08.cancer_calling_summary.R",
  "Scores the callers against each other from the tables alone.",
  "09", "09.cancer_calling_report.Rmd",
  "This report."
)
show_table(steps)
```

The scripts this was collated from, and what changed:

- `9.copykat.R` ran CopyKAT over the whole GEM well at once. That is wrong here, not merely
  superseded - the well pools four donors and CopyKAT's baseline comes from the cells it is
  given, so one donor's aneuploidy could define another's normal. `9a.copykat_perSample.R` is
  the per-donor version and is what produced the surviving outputs.
- the donor loop in `9a.copykat_perSample.R` was `unique(temp$sample_name)[c(4)]`, a hardcoded
  index, so it did one donor per invocation with the index edited by hand between runs. It now
  loops over all of them.
- `9b.numbat.sh` declared `SAMPLES` twice, LK1 first and LK2 second, so the second assignment
  silently overrode the first and the LK1 block was dead code that looked live. It also had
  `set -euo pipefail` commented out, so a failure inside the container did not fail the array
  task and a broken donor looked finished. Both fixed, and the run is now an argument.
- `10.numbat_annotation.R` (LK2) and `13.numbat_annotation.R` (LK1) are the same script at two
  stages. 13 hardcodes the donor directories with two commented out, and has its
  `readRDS` commented out so it only ran if the object was already in the session. 10 derives
  the directories, records which Numbat outputs each donor actually has, and handles
  incomplete runs. Step 04 is 10, parameterised by run.
- `12c.compare_numbat_copykat.R`, `12c.compare_numbat_copykat_501.R` and
  `12d.compare_numbat_copykat_noDoublets.R` are one script run three times with the donor
  edited in place. The entire difference between 12c and 12d is one `scDblFinder` subset call
  and a different output directory name, which is now a flag.
- `12d.investigate_clones.R` is **not** a separate analysis - 326 of its 333 non-comment lines
  are verbatim inside `12d_noDoublets`. It is the earlier standalone version, already folded
  in, so step 05 covers it.
- `20a.compare_all_clone_methods.R` is deliberately out of scope: it also brings in the
  mitochondrial clone callers and belongs with that report.

# 7. Data locations

```{r locations, results='asis'}
locs <- tibble::tribble(
  ~What, ~Where,
  "Annotated input, pilot", sprintf("results/seurat_annotated/%s/LK1_GEX_CITE_DSB_BoneMarrowMap_projected.rds", pilot),
  "Annotated input, LK2", sprintf("results/seurat_annotated/%s/demux_singlets_annotated_seurat.rds", run),
  "Cell Ranger BAM (Numbat input)", "results/cellranger_withbam/<run>/<sample>-GEX/outs/possorted_genome_bam.bam",
  "Phased SNP panel", "annotation/genome1K.phase3.SNP_AF5e2.chr1toX.hg38.sorted.vcf.gz",
  "Numbat container", "containers/numbat-rbase_latest.sif",
  "CopyKAT output", sprintf("results/seurat_annotated/%s/copykat_annotated/<donor>", pilot),
  "Numbat output, pilot", sprintf("results/seurat_annotated/%s/numbat/LK1_<donor>/numbat_final", pilot),
  "Numbat output, LK2", sprintf("results/seurat_annotated/%s/numbat/LK2_<donor>/numbat_final", run),
  "Numbat output, pipeline", sprintf("results_nf/%s/09c_numbat_run", run),
  "Long-read comparison", "results/longread_numbat/comparison_figures, comparison_tables",
  "Benchmark tables and figures (step 08)", "results/benchmarking/cancer_calling"
)
show_table(locs)
```

```{r tools, results='asis'}
tools <- tibble::tribble(
  ~Tool, ~Version, ~Use,
  "CopyKAT", "see sessionInfo", "expression-only CNV calling, per donor",
  "Numbat", "container `numbat-rbase_latest.sif` - unpinned `latest` tag, version not recorded",
  "allele-aware CNV and clone inference",
  "Eagle", "v2.4.1, inside the Numbat container", "phasing",
  "Seurat", "5.4.0", "object handling and plotting"
)
show_table(tools, "Tools")
```

# 8. Caveats

- steps 01-05 and 07 are the collated, cleaned versions of the scripts that produced the
  figures shown here; they were not re-executed for this report. Steps 06 and 08 were run
  fresh, which is what `Analysis finished` reflects
- the Numbat container is pinned to the `latest` tag, so the exact version that produced these
  results is not recorded and cannot be reconstructed from the image name. Any re-run may not
  reproduce them
- **Numbat completed for only two of the four pilot donors.** HBDN392-AML-MDS and MOLM13 are
  entirely `not_called`, so the caller comparison rests on two donors
- the CopyKAT/Numbat agreement is not an independent comparison in the lymphoid compartment,
  because those cells are CopyKAT's normal anchor. It is independent in the myeloid and
  progenitor compartments
- **HBDN206-MNpCT is called 81% aneuploid by CopyKAT and 74% tumour by Numbat**, and its name
  does not carry an AML designation the way the other pilot donors' do. Either it is a
  malignant sample that is simply not named as one, or both callers are failing on it in the
  same way. Nothing here settles which, and it should be checked against the sample sheet
  before these fractions are quoted
- the long-read comparison covers three LK2 donors only, and the donorN directories in the
  long-read output are positional - the mapping to sample names is an assumption, now asserted
  in step 07 rather than trusted silently
- step 06 reads the pipeline Numbat output under `results_nf/`, not the standalone output the
  other steps use. They are separate runs and their numbers are not interchangeable
