---
title: "LK1 re-sequenced"
subtitle: "`r params$run_id`"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  run_id: "260522_VH01624_461_222JLJVNX"
  prev_run: "260423_VH01624_453_222HWMYNX"
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  sample_name: "LK1-GEX"
---

<!--
# ------------------------------------------------------------------
# LK1 re-sequenced - step 05 of 05
#
# Renders the lab archive report from the outputs of steps 01-04. Tables come
# from the summary CSVs each step wrote, figures are rasterised out of the PDFs
# those steps produced.
#
# Render with ./render_report.sh, which also prints the HTML to PDF. pandoc is
# not on PATH on this laptop; it is inside the RStudio bundle.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})

theme_set(theme_bw(base_size = 12))

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

# Convert a PDF figure to PNG and emit an image tag, so figures render inline
# rather than being named as paths. Copied from 21.run_qc_report.Rmd.
embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) FALSE)
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

embed_image <- function(path, width_pct = 100) {
  if (is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', path, width_pct))
}

# Anything not on disk is skipped without comment.
show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  if (grepl("\\.pdf$", path, ignore.case = TRUE)) {
    embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
  } else embed_image(path, width_pct = width_pct)
}

`%||%` <- function(a, b) if (is.null(a)) b else a

proj <- params$proj
run  <- params$run_id
prev <- params$prev_run
res  <- file.path(proj, "results")
bm   <- file.path(res, "benchmarking", "depth_benchmarking")
bmd  <- file.path(res, "benchmarking", "depth_benchmarking_cellranger_downsampled")

read_if <- function(path, ...) {
  if (!file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

# Figures this report imports, resolved up front so the header can state when
# the last of them was produced.
FIGS <- c(
  file.path(bm, c("comparison_genes_vs_AML_atlas.pdf", "comparison_UMIs_vs_AML_atlas.pdf",
                  "median_genes_barplot.pdf", "median_UMIs_barplot.pdf",
                  "combined_benchmark_plot.pdf")),
  file.path(bmd, c("curve_cells_detected_vs_AML.pdf", "curve_genes_per_cell_vs_AML.pdf",
                   "curve_UMIs_per_gene_vs_AML.pdf", "curve_UMIs_per_cell_vs_AML.pdf",
                   "curve_reads_per_cell.pdf",
                   "violin_genes_UMIs_per_cell_full_depth_vs_AML_10x3v3.pdf",
                   "barplot_cell_counts_by_cell_type.pdf",
                   "boxplot_cells_per_donor_by_assay.pdf"))
)
FIGS <- FIGS[file.exists(FIGS)]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA
```

**Run:** `` `r run` `` (re-sequencing of `` `r prev` ``)

**Status:** current. This is the LK1 data used by all later analyses; the pilot run
`` `r prev` `` it replaces is superseded.

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `05.LK1_resequenced_report.Rmd`, from the outputs of
steps 01-04 in this folder.

LK1 was sequenced a second time to decide whether the pilot depth was sufficient, and whether
the rest of the cohort needed deeper sequencing. The comparison is against the pilot run and
against the published AML scAtlas.

For full report, download `LK1_resequenced_report.html` in the attachment and open in a
browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/LK1-resequenced>

# Aim

Identify whether we are getting a comparable amount of key features to the AML scAtlas -
cells, genes and UMIs - and evaluate whether further sequencing is required, via:

- subsampling of reads
- differential expression between cell types
- separation of clusters

# Conclusions

- AML scAtlas basic QC stats are comparable to our dataset, on average lower
- subsampling of reads shows that we have reached saturation levels for cell detection
- we have **not** reached saturation levels for: number of genes detected; number of UMIs per
  gene; number of DE genes for cells with either low (T-cells) or high (HSCs) numbers of genes
  expressed
- cluster separation of HSCs from other precursors (MEP) is slightly better for re-sequenced
  data
- re-sequencing gained about 25% more genes and 39% more UMIs per cell for essentially no
  extra cells, which is the expected shape when a library is under-sequenced rather than
  under-loaded

# 1. What re-sequencing gained

```{r gain, results='asis'}
show_table(read_if(file.path(bm, "first_run_vs_resequenced_gain_summary.csv")),
           caption = "First run vs re-sequenced, per-cell gain")
show_table(read_if(file.path(bm, "benchmark_summary.csv")),
           caption = "Our two runs against the AML scAtlas")
```

Cell numbers barely move while genes and UMIs per cell rise by a quarter and a third
respectively. That is the signature of a library that was sequenced too shallowly rather than
loaded too lightly - more reads recover more of the transcriptome from the cells already
captured, rather than finding new cells.

Both runs sit **above** the atlas medians on genes and UMIs per cell, which is what "atlas
stats are on average lower" in the conclusions refers to.

```{r saturation, results='asis'}
m <- read_if(file.path(bm, "cellranger_metrics_summary_combined.csv"))
if (!is.null(m)) {
  keep <- intersect(c("Run", "Sample ID", "Estimated Number of Cells", "Mean Reads per Cell",
                      "Median Genes per Cell", "Number of Reads", "Sequencing Saturation"),
                    names(m))
  show_table(m[, keep, drop = FALSE], caption = "Cell Ranger metrics, both runs")
}
```

Sequencing saturation stays in the mid-40s in both runs - consistent with the curves below,
where cell detection has plateaued but per-cell content has not.

```{r figs-gain, results='asis'}
show_fig(file.path(bm, "median_genes_barplot.pdf"), "Median genes per cell, first run vs re-sequenced.")
show_fig(file.path(bm, "median_UMIs_barplot.pdf"), "Median UMIs per cell, first run vs re-sequenced.")
show_fig(file.path(bm, "combined_benchmark_plot.pdf"), "Combined benchmark across both runs and the atlas.")
```

# 2. Against the AML scAtlas

```{r figs-atlas, results='asis'}
show_fig(file.path(bm, "comparison_genes_vs_AML_atlas.pdf"),
  "Genes per cell, our runs against the atlas.")
show_fig(file.path(bm, "comparison_UMIs_vs_AML_atlas.pdf"),
  "UMIs per cell, our runs against the atlas.")
show_fig(file.path(bmd, "violin_genes_UMIs_per_cell_full_depth_vs_AML_10x3v3.pdf"),
  paste("Restricted to 10X 3' v3, the assay that matches ours - the fair comparison, since",
        "the atlas pools several chemistries with different capture efficiency."))
show_fig(file.path(bmd, "barplot_cell_counts_by_cell_type.pdf"),
  "Atlas cell counts by cell type, for context on what the reference contains.")
show_fig(file.path(bmd, "boxplot_cells_per_donor_by_assay.pdf"),
  "Cells per donor by assay in the atlas.")
```

# 3. Saturation curves

Cell Ranger's molecule-level downsampling, re-analysed at each depth and plotted against the
atlas. These are the figures the saturation conclusions rest on.

```{r figs-curves, results='asis'}
show_fig(file.path(bmd, "curve_cells_detected_vs_AML.pdf"),
  "**Cells detected vs depth - this one has plateaued.** Extra reads do not recover extra cells.")
show_fig(file.path(bmd, "curve_genes_per_cell_vs_AML.pdf"),
  "**Genes per cell vs depth - still climbing.** No plateau, so more depth would still add genes.")
show_fig(file.path(bmd, "curve_UMIs_per_gene_vs_AML.pdf"),
  "**UMIs per gene vs depth - still climbing.**")
show_fig(file.path(bmd, "curve_UMIs_per_cell_vs_AML.pdf"), "UMIs per cell vs depth.")
show_fig(file.path(bmd, "curve_reads_per_cell.pdf"), "Reads per cell across the depth series.")
```

# 4. Effect of depth on downstream analysis

Two things were checked beyond raw counts: whether differential expression between cell types
holds up at lower depth, and whether clusters still separate.

Cluster counts across the depth series, from the analysis in step 07 (reproduced from the
read-depth deck, as the figure itself is not among the saved outputs):

| Depth | Total clusters | HSC clusters |
|---|---|---|
| 10% subsample | 5 | 2 |
| 30% subsample | 8 | 3 |
| 60% subsample (~pilot 1) | 8 | 4 |
| Full dataset | 9 | 4 |

The full dataset resolves one cluster the 60% subsample does not - naive T-cells - and HSC
separation from other precursors improves with depth. The pilot run sits at roughly the 60%
point of the re-sequenced data, which is why its HSC/MEP separation is the weaker of the two.

# 5. Data locations

| What | Location |
|---|---|
| Sequencing facility output | `` `r file.path("/pipeline/Runs/NextSeq", run, "ProjectFolders/Project_Lev-Kats")` `` |
| Working copy, cluster | `` `r file.path("/scratch/users/nbartonicek/projects/amgen/raw_reseq", run)` `` - `Sample_LK1-GEX/` |
| Cell Ranger output | `` `r file.path("/scratch/users/nbartonicek/projects/amgen/results/cellranger_withbam", run)` `` |
| Depth benchmarking | `results/benchmarking/depth_benchmarking/` and `.../depth_benchmarking_cellranger_downsampled/` |
| Reference | Cell Ranger annotation `refdata-gex-GRCh38-2024-A` |

```{r out-dirs, results='asis'}
# Shallow listing only. cellranger_withbam holds ~6,300 files for this run and
# stat-ing all of them over the SMB mount takes longer than the whole render,
# so this walks the immediate children and uses their timestamps.
dirs <- c("cellranger_withbam", "cite_seq_count", "emptydrops", "genotype_demux",
          "scDblFinder", "seurat_demux", "cellbender")
rows <- lapply(dirs, function(d) {
  p <- file.path(res, d, run)
  if (!dir.exists(p)) return(NULL)
  e <- list.files(p, full.names = TRUE, all.files = FALSE)
  e <- e[!grepl("(^|/)\\._", e)]
  if (!length(e)) return(NULL)
  dt <- as.Date(file.info(e)$mtime)
  tibble(directory = d, entries = length(e),
         first = min(dt, na.rm = TRUE), last = max(dt, na.rm = TRUE))
})
show_table(bind_rows(rows),
           caption = paste0("Output directories under results/<dir>/", run,
                            " (top-level entries)"))
```

The re-sequenced run was then taken through the standard processing - emptyDrops,
CITE-seq-Count, cellSNP/vireo, scDblFinder, CellBender. Those steps are benchmarked in their
own reports rather than here; this one covers the re-sequencing decision only.

# 6. What was run, in order

| Step | Script | What it does |
|---|---|---|
| 01 | `01.cellranger_resequence.sh` | Cell Ranger 10.0.0 count on the re-sequenced LK1 GEX library. |
| 02 | `02.investigate_resequencing.R` | First run vs re-sequenced: cells, genes, UMIs, and the gain summary. |
| 03 | `03.subsampling_vs_atlas.R` | Our depth against the AML scAtlas, plus the molecule-level saturation curves. |
| 04 | `04.DE_subsampling.R` | Pseudobulk DE at each depth, then cluster HBDN501-AML-KMT2A and score separation. |
| 05 | `05.LK1_resequenced_report.Rmd` | This report. |

Each analysis appears once. Where a script went through successive revisions, the version
kept is the one that produced the outputs shown, and its header names the revisions it
supersedes:

| Step | Kept | Superseded |
|---|---|---|
| 03 | `1c1...` (2026-06-01) | `1c...` (2026-05-27, first pass). A later `1c2...` (2026-06-16) also exists but produced no output - nothing under `depth_benchmarking*` postdates 2026-06-01 - and is *shorter* than the kept version, so it is not a strict superset. |
| 04 | `1d2...` (2026-06-01) | `1d...` (2026-05-28, first pass) and `1d3...DE_subsampling_LK2.R`, which is byte-identical despite the `_LK2` name and analyses LK1. |

# 7. Tools

| Tool | Version | Source |
|---|---|---|
| Cell Ranger | 10.0.0 | `~/tools/cellranger-10.0.0/cellranger` - the re-sequencing needed this version |
| Cell Ranger annotation | `refdata-gex-GRCh38-2024-A` | `annotation/` |
| AML scAtlas | published reference | `results/AML_atlas/` |

```{r session}
sessionInfo()
```

# 8. Caveats

1. **The depth series here is Cell Ranger's molecule-level downsampling**, not a FASTQ-level
   re-run. The seqtk FASTQ downsampling with `frac_000` to `frac_090` directories is a
   separate, later exercise carried out on LK2 (`cellranger_downsampled_reseq/260528…`,
   2026-06-15) and does not belong to this report.
2. **Step 05 hardcodes a downsampled path under this run** that does not exist
   (`cellranger_downsampled_reseq/260522…`); only the LK2 equivalent is on disk. It runs
   against the molecule-level outputs regardless, but the path should be corrected before the
   script is reused.
3. The cluster-count table in section 4 is taken from the read-depth deck; the figure behind
   it is not among the saved outputs.
