---
title: "Benchmarking: CITE-seq analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  pilot_run: "260423_VH01624_453_222HWMYNX"
  run: "260528_VH01624_464_222K7VKNX"
---

<!--
# ------------------------------------------------------------------
# CITE-seq benchmark - step 07 of 07
#
# Renders the lab archive report from the outputs of steps 01-06.
# Render with ./render_report.sh, which also prints the HTML to PDF.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))
options(scipen = 999)

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) {
    cat(sprintf("\n_(could not rasterise `%s`: %s)_\n\n",
                basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

read_if <- function(path, ...) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj  <- params$proj
res   <- file.path(proj, "results")
run   <- params$run
pilot <- params$pilot_run

dsb_dir   <- file.path(res, "cite_seq_dsb")
qc_dir    <- file.path(res, "benchmarking", "cite_qc_then_dsb")
igg_dir   <- file.path(res, "benchmarking", "dsb_igg_subsample")
unmap_dir <- file.path(res, "adt_unmapped_diagnostics")
unmap_inv <- file.path(res, "adt_unmapped_investigator")
cmp_dir   <- file.path(res, "seurat_annotated")

find_one <- function(root, name) {
  if (is.null(root) || !dir.exists(root)) return(NA_character_)
  hits <- list.files(root, pattern = paste0("^", gsub("\\.", "\\\\.", name), "$"),
                     recursive = TRUE, full.names = TRUE)
  hits <- hits[!grepl("(^|/)\\._", hits)]
  if (!length(hits)) NA_character_ else hits[1]
}

all_pdfs <- function(root) {
  if (!dir.exists(root)) return(character(0))
  f <- list.files(root, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE)
  f[!grepl("(^|/)\\._", f)]
}

FIGS <- unique(c(all_pdfs(dsb_dir), all_pdfs(qc_dir), all_pdfs(igg_dir),
                 all_pdfs(unmap_dir),
                 find_one(cmp_dir, "01_marker_mean_shift_DSB_minus_CLR.pdf")))
FIGS <- FIGS[!is.na(FIGS) & file.exists(FIGS)]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA
```

**Runs:** pilot `` `r pilot` `` for the first DSB evaluation and the CLR-vs-DSB comparison,
then `` `r run` `` (LK2) for the production CITE QC.

**Status:** settled. DSB is the normalisation the project uses; CLR is kept only as the
comparison baseline.

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `07.cite_seq_report.Rmd`, from the outputs of steps
01-06 in this folder.

ADT counts arrive from CITE-seq-Count as raw tag counts per barcode, and are dominated by
ambient antibody. Two normalisations were compared: CLR, which scales within each cell, and
DSB, which subtracts a background estimated from empty droplets and then removes per-cell
technical noise using the isotype controls.

For full report, download `cite_seq_report.html` in the attachment and open in a browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/5.Benchmarking-CITE-seq_analysis>

# Aim

Decide how to normalise the ADT signal, and confirm the counting step is not losing reads:

- does DSB give more specific markers than CLR
- is the background estimate stable, or an artefact of which isotype controls are used
- how many ADT reads come back unmapped, and whether the tag reference and read structure
  are correct

# Conclusions

- **DSB gives increased marker specificity over CLR**, scored as mean of the expected
  population divided by median background, and is what the project uses
- the background estimate is stable: 30x random subsampling of the isotype controls does not
  move it materially
- populations are called at **3 SD over background**, at both the broad and the granular
  cell type level
- noise is estimated from ten isotype controls - IgG_Fc, Mouse IgG1, IgG2a and IgG2b, Rat
  IgG2b, IgG1_k, IgG1_l, IgG2a_k and IgG2c_k, and Armenian Hamster IgG
- mapping is not a problem: CITE-seq-Count maps **97% of ADT reads** and 95% of HTO reads
- the unmapped remainder is **poly-G** - 90 bases of G, 99.9% of the top-100 unmapped reads on
  the pilot ADT and 98.9% on HTO. That is the two-colour chemistry artefact, where a dark or
  failed cluster is base-called as G. These are not antibody tags, so they are neither a
  reference problem nor reads from discarded droplets
- step 06 as run is a whole-library check rather than an unmapped-read diagnosis, and the
  library passes it: the abundant sequences are exact matches to the reference in the correct
  orientation

# 1. Counting the tags

Step 01 runs CITE-seq-Count against the emptyDrops barcode whitelist, so the ADT matrix
covers the same barcodes as the RNA. Step 02 is the same command with `-u` and `-ut 100000`
added, which writes the unmapped tags out instead of discarding them - that table is what
step 06 diagnoses. Both are kept because the production run does not produce the unmapped
output, and the diagnostic run was done separately in July.

Note that the ADT is counted **outside** Cell Ranger: the Cell Ranger matrix for these runs
contains `Gene Expression` features only. Anything that operates on the Cell Ranger matrix,
CellBender included, therefore never sees the protein data.

# 2. CITE QC and DSB

Step 03 is the main workflow: load the demultiplexed object, import ADT counts for the
emptyDrops cells, do the CLR-based QC first, then select background droplets from the raw
output and run DSB with matched plots so the two can be compared like for like.

```{r qc-figs, results='asis'}
qnotes <- list(
  "03_cell_QC_RNA_ADT_raw.pdf" = "Cell-level QC on RNA and raw ADT together.",
  "04_raw_ADT_marker_distributions.pdf" = "Raw ADT counts per marker, before any normalisation - the ambient signal is what makes these unusable as they stand.",
  "05_CLR_ADT_marker_distributions.pdf" = "The same markers after CLR.",
  "06_RNA_umap_QC.pdf" = "RNA UMAP used for the overlays below.",
  "07_CLR_ADT_featureplots_top_markers.pdf" = "Top markers under CLR on the RNA UMAP.",
  "08_CLR_ADT_discriminative_markers.pdf" = "The most discriminative markers under CLR.",
  "09_CLR_heatmap_by_RNA_cluster.pdf" = "CLR signal by RNA cluster.",
  "11_CITE_marker_CLR_positive_cells_by_sample.pdf" = "Positive cells per marker per sample under CLR."
)
for (nm in names(qnotes)) show_fig(find_one(dsb_dir, nm), qnotes[[nm]])
```

# 3. Background and the isotype controls

DSB estimates background from empty droplets and then removes per-cell technical noise using
the isotype controls. Step 04 checks that estimate is not an artefact of which isotypes happen
to be in the panel, by subsampling them 30 times at random and re-deriving the background.

```{r igg-figs, results='asis'}
show_fig(find_one(igg_dir, "dsb_igg_reduction_subsampling.pdf"),
  paste("Background estimate across 30 random subsamples of the isotype controls. A tight",
        "spread means the estimate does not depend on which controls are included, which is",
        "what justifies using it per cell."))
```

# 4. DSB against CLR

Step 05 is the head-to-head, on HBDN206-MNpCT.

```{r cmp-figs, results='asis'}
show_fig(find_one(cmp_dir, "01_marker_mean_shift_DSB_minus_CLR.pdf"),
  paste("Per-marker mean shift, DSB minus CLR. Markers above zero gained signal relative to",
        "background under DSB."))
```

```{r cmp-tables, results='asis'}
for (nm in c("01_marker_mean_shift_DSB_minus_CLR.csv",
             "02_RNA_protein_correlation_DSB_vs_CLR.csv",
             "03_marker_celltype_specificity_DSB_vs_CLR.csv",
             "07_expected_marker_specificity_DSB_vs_CLR.csv")) {
  show_table(read_if(find_one(cmp_dir, nm)), caption = nm, n = 25)
}
```

The specificity score behind the conclusion is **mean of the expected population divided by
median background**, computed per marker against a hand-curated panel: CD163, CD14, CD33 and
CD11b for myeloid/DC; CD34, CD117 and CD133 for stem/progenitor; CD4, CD8, CD19 and CD20 for
lymphoid; CD71 for erythroid. Cells are called positive at 3 SD over background.

# 5. Unmapped ADT reads

**Read this section for what it is.** Step 06 was run with `input_mode <- "full_fastq"`
against the entire LK2 ADT R2 file, not against the unmapped reads. Everything below therefore
describes the whole library, and is a check that the tag reference and read structure are
right - which they are. It is not a diagnosis of the unmapped reads.

The unmapped reads are separately accounted for. CITE-seq-Count's own `run_report.yaml` puts
mapping at **97%** for ADT and **95%** for HTO, and its `unmapped.csv` shows what the
remainder is:

| Run / library | Poly-G share of the top-100 unmapped |
|---|---|
| Pilot ADT (`260423`, LK1) | 99.9% |
| Pilot HTO (`260423`, LK1) | 98.9% |
| LK2 ADT (`260528`) | 67.5% |

Those sequences are 90 bases of `G`. On two-colour chemistry a dark or failed cluster is
base-called as G, so this is empty flowcell rather than anything about the antibodies, the
reference or the droplets. `unmapped.csv` lists only the top 100 sequences, so this
characterises the dominant species rather than every unmapped read - and the LK2 figure of
67.5% means a third of its unmapped reads are something else, which has not been looked at.

```{r unmapped-figs, results='asis'}
show_fig(find_one(unmap_dir, "edit_distance_histogram.pdf"),
  paste("Edit distance from each observed barcode to the nearest known tag. A pile-up at",
        "distance 1-2 means sequencing error; a pile-up further out means the observed",
        "sequences are not the tags at all."))
show_fig(find_one(unmap_dir, "top_observed_barcodes.pdf"),
  "The most frequent observed barcodes among the unmapped reads.")
```

```{r unmapped-tables, results='asis'}
for (nm in c("observed_barcode_counts.csv", "observed_vs_known_distance.csv",
             "offset_scan_summary.csv", "reverse_complement_distance.csv")) {
  show_table(read_if(find_one(unmap_inv, nm)), caption = nm, n = 20)
}
```

The four outputs test four explanations in turn: whether the observed sequences resemble the
known tags at all, how far they are from them, whether the tag sits at a different offset in
the read than expected, and whether it has been reverse complemented. Across the top 1,000
observed sequences, covering 96% of reads, 93 percentage points are exact matches to a real
antibody and 96 are within two edits, while the reverse-complement check sits at distance 5-7
with no matches. The library and reference are sound.

To diagnose the unmapped reads properly, step 06 would be re-run with
`input_mode <- "unmapped_fastq"` against the unmapped output from step 02
(`cite_seq_count/<run>/unmapped_tags_<sample>.csv`).

# 6. What was run, in order

| Step | Script | What it does |
|---|---|---|
| 01 | `01.cite_seq_count.sbatch` | CITE-seq-Count on the ADT library, emptyDrops whitelist. Production counting. |
| 02 | `02.cite_seq_count_unmapped.sbatch` | Same run with `-u` / `-ut 100000` so unmapped tags are written out. |
| 03 | `03.CITE_qc.R` | CITE QC then DSB: raw totals, CLR, marker distributions, overlays, heatmaps, background selection, DSB. |
| 04 | `04.evaluate_DSB.R` | DSB against background droplets with 30x isotype subsampling. |
| 05 | `05.compare_DSB_vs_CLR.R` | Head-to-head DSB vs CLR on one sample. |
| 06 | `06.CITE_unmapped_QC.R` | Diagnoses the unmapped ADT reads. |
| 07 | `07.cite_seq_report.Rmd` | This report. |

`scripts/7d.DSB_cluster_resolution.R` is not a separate step: it is step 05 re-pointed at LK2,
differing by twelve lines that are all run and sample variables. Change `run`, `sample_name`
and `sample_short` in step 05 to reproduce it. Its name is misleading - it is not a
cluster-resolution analysis.

# 7. Data locations

| What | Location |
|---|---|
| ADT counts | `results/cite_seq_count/<run>/adt_counts_*/` |
| CITE QC and DSB | `results/cite_seq_dsb/<run>/` |
| Earlier QC-then-DSB outputs | `results/benchmarking/cite_qc_then_dsb/` |
| Isotype subsampling | `results/benchmarking/dsb_igg_subsample/` |
| DSB vs CLR | `results/seurat_annotated/<run>/projectionFigures/DSB_vs_CLR_<donor>/` |
| Unmapped diagnostics | `results/adt_unmapped_diagnostics/` and `results/adt_unmapped_investigator/` |

# 8. Tools

| Tool | Version | Source |
|---|---|---|
| CITE-seq-Count | to confirm | steps 01 and 02 |
| dsb | see session below | denoised and scaled by background |
| Seurat CLR | see session below | comparison baseline |

```{r session}
sessionInfo()
```

# 9. Caveats

1. **The residual CITE analysis is not included.** `scripts/13.CITE_residuals.R` belongs to this
   period but cannot be run as written - it never loads an object, expecting `seu_dsb` and
   `celltype_cols` to exist in the session already - and its output directory
   `results/benchmarking/residual_cite_analysis` is empty. It will be added once it has been
   run.
2. **The CLR-vs-DSB comparison is one donor**, HBDN206-MNpCT on the pilot run. The LK2
   equivalent exists only as the re-pointed script noted above.
3. **CellBender does not touch the protein data** in the current pipeline, since the Cell
   Ranger matrix is `Gene Expression` only. Whether ambient protein removal before DSB helps
   is a separate question, and is not tested here.
