---
title: "Benchmarking: cell type annotation"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
params:
  proj: "/Volumes/bioinf_scratch/users/nbartonicek/projects/amgen"
  pilot_run: "260423_VH01624_453_222HWMYNX"
  run: "260528_VH01624_464_222K7VKNX"
---

<!--
# ------------------------------------------------------------------
# Cell type annotation benchmark - step 05 of 05
#
# Renders the lab archive report from the outputs of steps 01-04.
# Render with ./render_report.sh, which also prints the HTML to PDF.
# ------------------------------------------------------------------
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 9)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(purrr); library(knitr); library(ggplot2); library(scales)
})
theme_set(theme_bw(base_size = 12))
options(scipen = 999)

HAS_PDFTOOLS <- requireNamespace("pdftools", quietly = TRUE)

embed_pdf_figure <- function(pdf_path, dpi = 120, page = 1, width_pct = 100) {
  if (is.na(pdf_path) || !file.exists(pdf_path)) return(invisible(NULL))
  if (!HAS_PDFTOOLS) return(invisible(NULL))
  png_path <- tempfile(fileext = ".png")
  ok <- tryCatch({
    pdftools::pdf_convert(pdf_path, format = "png", pages = page, dpi = dpi,
                          filenames = png_path, verbose = FALSE); TRUE
  }, error = function(e) {
    cat(sprintf("\n_(could not rasterise `%s`: %s)_\n\n",
                basename(pdf_path), conditionMessage(e)))
    FALSE
  })
  if (ok) cat(sprintf('\n<img src="%s" style="width:%d%%"/>\n\n', png_path, width_pct))
}

show_fig <- function(path, note = NULL, width_pct = 100, dpi = 120) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(invisible(NULL))
  cat(sprintf("\n**`%s`**\n\n", basename(path)))
  if (!is.null(note)) cat(note, "\n\n")
  embed_pdf_figure(path, dpi = dpi, width_pct = width_pct)
}

read_if <- function(path, ...) {
  if (is.null(path) || is.na(path) || !file.exists(path)) return(NULL)
  out <- try(suppressMessages(readr::read_delim(path, show_col_types = FALSE, ...)), silent = TRUE)
  if (inherits(out, "try-error")) NULL else out
}

show_table <- function(df, caption = NULL, n = Inf) {
  if (is.null(df) || nrow(df) == 0) return(invisible(NULL))
  if (is.finite(n)) df <- head(df, n)
  cat("\n"); print(kable(df, caption = caption)); cat("\n\n")
}

proj  <- params$proj
res   <- file.path(proj, "results")
run   <- params$run
pilot <- params$pilot_run

bench_dir <- file.path(res, "benchmarking", "celltype_annotation")
pilot_dir <- file.path(res, "seurat_annotated", pilot)
lk2_dir   <- file.path(res, "seurat_annotated", run)
unmap_dir <- file.path(lk2_dir, "LK2-GEX_results_nf_unannotated_check")

find_one <- function(root, name) {
  if (is.null(root) || !dir.exists(root)) return(NA_character_)
  hits <- list.files(root, pattern = paste0("^", gsub("\\.", "\\\\.", name), "$"),
                     recursive = TRUE, full.names = TRUE)
  hits <- hits[!grepl("(^|/)\\._", hits)]
  if (!length(hits)) NA_character_ else hits[1]
}

all_pdfs <- function(root) {
  if (!dir.exists(root)) return(character(0))
  f <- list.files(root, pattern = "\\.pdf$", recursive = TRUE, full.names = TRUE)
  f[!grepl("(^|/)\\._", f)]
}

FIGS <- unique(c(all_pdfs(bench_dir), all_pdfs(unmap_dir),
                 all_pdfs(file.path(pilot_dir, "projectionFigures")),
                 find_one(pilot_dir, "00a_BoneMarrowMap_reference_broad_celltype.pdf"),
                 find_one(lk2_dir, "01_mapping_error_QC.pdf")))
FIGS <- FIGS[!is.na(FIGS) & file.exists(FIGS)]
FIGS <- FIGS[file.info(FIGS)$size > 0]
analysis_finished <- if (length(FIGS)) max(as.Date(file.info(FIGS)$mtime)) else NA

# tables written by step 04
t_qc_run   <- read_if(file.path(bench_dir, "01_mapping_QC_by_run.csv"))
t_qc_samp  <- read_if(file.path(bench_dir, "02_mapping_QC_fail_by_sample.csv"))
t_profile  <- read_if(file.path(bench_dir, "03_rejected_cell_profile.csv"))
t_conf     <- read_if(file.path(bench_dir, "04_label_confidence.csv"))
t_conf_ct  <- read_if(file.path(bench_dir, "05_label_confidence_by_celltype.csv"))
t_relabel  <- read_if(file.path(bench_dir, "06_initial_vs_final_label.csv"))
t_vocab    <- read_if(file.path(bench_dir, "07_fine_vs_broad_vocabulary.csv"))
t_map      <- read_if(file.path(bench_dir, "09_fine_to_broad_label_map.csv"))
t_thresh   <- read_if(file.path(bench_dir, "10_MAD_threshold_cost.csv"))

# tables written by step 03
u_qc       <- read_if(file.path(unmap_dir, "unknown_vs_annotated_qc_medians.csv"))
u_patient  <- read_if(file.path(unmap_dir, "unknown_fraction_by_patient.csv"))
u_hto      <- read_if(file.path(unmap_dir, "unknown_vs_annotated_HTO_classification.csv"))
u_nn_nat   <- read_if(file.path(unmap_dir, "unknown_nearest_neighbor_identity_umap.csv"))
u_nn_proj  <- read_if(file.path(unmap_dir, "unknown_nearest_neighbor_identity_umap_projected.csv"))
u_cluster  <- read_if(file.path(unmap_dir, "unknown_cells_by_seurat_cluster.csv"))

pct <- function(tbl, col, ...) {
  if (is.null(tbl)) return("n/a")
  v <- tbl %>% filter(...) %>% pull({{ col }})
  if (!length(v)) "n/a" else as.character(v[1])
}
```

**Runs:** pilot `` `r pilot` `` (HBDN206-MNpCT, HBDN392-AML-MDS, HBDN501-AML-KMT2A,
MOLM13) and `` `r run` `` / LK2 (normal-01, HBDN376-TP53, HBDN498-TP53, APOP576-TP53).

**Analysis finished:** `r if (is.na(analysis_finished)) "n/a" else format(analysis_finished)` -
the most recent of the `r length(FIGS)` figures shown below.

**Report generated:** `r Sys.Date()` by `05.celltype_annotation_report.Rmd`, from the outputs
of steps 01-04 in this folder.

Cells are annotated by projecting them into the Zeng et al. bone marrow reference with
Symphony, rather than by clustering and scoring marker genes. The reference carries its own
mapping-error QC, so a cell can come back with no label at all. This report covers how the
projection behaves on our data, how many cells it refuses, what those cells are, and how much
the labels can be trusted.

For full report, download `celltype_annotation_report.html` in the attachment and open in a
browser.

Code used - attached and in the github repository:
<https://github.com/nbartonicek/surface_proteome_dev/tree/main/lab_archives/7.Benchmarking-cell_type_annotation>

# Aim

Establish reference-based annotation as the project's cell typing, and find out where it
breaks:

- how many cells does the reference refuse to label, and is that number stable between runs
- what are the refused cells - real biology the reference does not contain, or bad cells
- how much confidence sits behind the labels that do survive, and for which populations
- fine versus broad labels: which one downstream analysis should key on

# Conclusions

- **BoneMarrowMap projection works and is what the project uses.** It annotates
  `r pct(t_qc_run, percent, run == "LK1 pilot", mapping_error_QC == "Pass")`% of pilot cells
  and `r pct(t_qc_run, percent, run == "LK2", mapping_error_QC == "Pass")`% of LK2 cells, at
  the default `MAD_threshold` of 2.5
- **the cells it refuses are bad cells, not unknown biology.** They are half the depth of the
  cells that pass (`r pct(t_profile, nFeature_RNA, run == "LK2", mapping_error_QC == "Fail")`
  vs `r pct(t_profile, nFeature_RNA, run == "LK2", mapping_error_QC == "Pass")` genes in LK2),
  have higher mitochondrial and **four times lower ribosomal** content, and are enriched for
  HTO-negative droplets. Forced to pick a neighbour they land on late erythroid and monocyte,
  the two lowest-complexity populations
- **the rejection rate is a readout of how far a sample sits from normal marrow.** In the pilot
  the normal donor loses 4.8% and the leukaemias 10-11%; in LK2 APOP576-TP53 loses **30%**
  against 3-4% for the rest. A high reject rate is a flag to look at the sample, not at the
  reference
- **`predict_CellTypes()` does not re-label anything.** The final label is identical to
  `initial_CellType` for 100% of cells; the only thing it does is set the QC-failed cells to
  NA. The NA cells and the `mapping_error_QC == "Fail"` cells are the same cells, 1:1
- **label confidence tracks how continuous the biology is, not how well the projection ran.**
  Terminal populations - Monocyte, cDC, pDC, Plasma Cell, Late Erythroid - come back at a
  median vote fraction of 1.0. HSC MPP, LMPP, Early Lymphoid and Cycling Progenitor sit at
  0.57-0.67, because the reference is resolving a gradient there rather than a boundary
- **downstream grouping must key on `predicted_CellType_Broad`, not on `predicted_CellType`.**
  The fine column has 53 levels and only 9 of them appear in the broad vocabulary the lineage
  lists and the bespoke palette are written against. Keyed on the fine label,
  `r pct(t_vocab, percent_cells_Other_if_keyed_on_fine, run == "LK1 pilot")`% of pilot cells
  and `r pct(t_vocab, percent_cells_Other_if_keyed_on_fine, run == "LK2")`% of LK2 cells fall
  into "Other"; keyed on the broad label, none do
- **the Unknown cells sit with erythroid and pro-monocyte.** On
  `02a_umap_native_unknown_highlighted.pdf` the majority fall in the erythroid / late
  erythroid territory and in pro-monocyte, rather than scattering across the map
- **the pipeline and the standalone scripts disagree on the threshold.**
  `nextflow/bin/annotate.R` runs at `MAD_threshold = 4`; steps 01 and 02 ran at the
  BoneMarrowMap default of 2.5. Going 2.5 -> 3 -> 4 roughly halves the rejection rate
  (pilot 7.95 -> 6.36 -> 3.91%, LK2 5.48 -> 3.58 -> 1.75%) without changing which samples are
  hardest to place. Given what section 3 shows the rejected cells to be, 2.5 is the better
  setting - but 4 is what production currently runs, and that should be settled on purpose
- there is **no confidence cutoff anywhere in the annotation**. `predict_CellTypes()` reports
  the k = 30 vote share and never thresholds on it; the only gate is the mapping-error QC

# 1. The reference

The reference is BoneMarrowMap (Zeng et al.), a Symphony reference built on normal human bone
marrow. It ships as a 345 MB reference object plus a 221 MB uwot model, both downloaded once
into `annotation/`. Projection is `map_Query()` -> `calculate_MappingError()` ->
`predict_CellTypes()` -> `predict_Pseudotime()`, with the demultiplexed sample as the batch
variable so Symphony harmonises across donors within a pool rather than treating the GEM well
as one batch.

```{r reference, results='asis'}
show_fig(find_one(pilot_dir, "00a_BoneMarrowMap_reference_broad_celltype.pdf"),
  paste("The reference itself, coloured by broad cell type in the project palette. Query",
        "cells are placed into this space; they do not move it. Stromal is the grey island -",
        "it is in the reference, and it is the population that used to be dropped from every",
        "figure by the palette bug described in section 9."), dpi = 90)
```

The reference returns two label columns, and the difference between them is the single thing
most likely to cause trouble downstream:

```{r vocab-map, results='asis'}
if (!is.null(t_map)) {
  ex <- t_map %>%
    group_by(predicted_CellType_Broad) %>%
    summarise(n_fine = n(),
              fine_labels = paste(head(predicted_CellType, 4), collapse = ", "),
              .groups = "drop") %>%
    arrange(desc(n_fine))
  show_table(head(ex, 10),
             "How the fine labels roll up into the broad ones (10 broadest, LK1 + LK2 pooled)")
}
```

# 2. How many cells the reference refuses

`calculate_MappingError()` scores each cell on how well it sits in reference space and calls
anything beyond `MAD_threshold` median absolute deviations a Fail. Failed cells keep their
projected coordinates but lose their label.

```{r qc-tables, results='asis'}
show_table(t_qc_run, "Mapping-error QC, whole run (MAD_threshold = 2.5)")
show_table(t_qc_samp, "Rejection rate per demultiplexed sample, worst first")
```

```{r qc-figs, results='asis'}
show_fig(file.path(bench_dir, "01_mapping_QC_fail_by_sample.pdf"),
  paste("Per-sample rejection. The normal donors sit at the bottom; the leukaemias and the",
        "MOLM13 cell line sit at the top. APOP576-TP53 is the outlier at 30%."))
show_fig(find_one(file.path(pilot_dir, "projectionFigures"), "01_mapping_error_QC.pdf"),
  "The mapping error distribution and the threshold, pilot run.")
show_fig(find_one(lk2_dir, "01_mapping_error_QC.pdf"),
  "The same for LK2.")
```

The ordering is the point. A reference built on normal marrow places normal cells well and
leukaemic cells less well, so the rejection rate reads as distance from normal rather than as
a technical failure. It is worth checking whenever a new sample comes in.

# 3. What the refused cells are

```{r profile, results='asis'}
show_table(t_profile, "Median QC covariates, cells that pass versus cells that fail")
show_fig(file.path(bench_dir, "02_rejected_cell_profile.pdf"),
  paste("The rejected cells are shallower, more mitochondrial and much less ribosomal than",
        "the cells that pass. The ribosomal drop is the clearest separator - roughly a",
        "four-fold difference in both runs."))
```

Step 03 pushes this further on the LK2 pipeline checkpoint, where the same projection is run
at `MAD_threshold = 4` and the failed cells are followed individually.

```{r unmapped-tables, results='asis'}
show_table(u_qc, "Unknown versus annotated cells, median QC covariates (MAD_threshold = 4)")
show_table(u_patient %>% dplyr::filter(is_unknown) %>% arrange(desc(percent)),
           "Unknown fraction by donor")
show_table(u_hto, "HTO classification of Unknown versus annotated cells")
```

Unknown cells are enriched in HTO-negative droplets
(`r if (!is.null(u_hto)) round(u_hto$percent[u_hto$is_unknown & u_hto$HTO_classification.global == "Negative"], 1) else "n/a"`%
against
`r if (!is.null(u_hto)) round(u_hto$percent[!u_hto$is_unknown & u_hto$HTO_classification.global == "Negative"], 1) else "n/a"`%),
which is the same population that fails hashing - low-content droplets that carry neither
enough transcriptome to place nor enough hashtag to assign.

```{r unmapped-figs, results='asis'}
unotes <- list(
  "01_unknown_fraction_by_cluster.pdf" = paste(
    "Unknown cells are not spread evenly: one native cluster (11) is a fifth Unknown, the",
    "rest are under 3%. They form a compartment, not a background."),
  "02_umap_native_with_unknown.pdf" = "Native UMAP with Unknown as its own category.",
  "02a_umap_native_unknown_highlighted.pdf" = paste(
    "The same, with only the Unknown cells picked out - they sit together rather than",
    "scattering across the manifold."),
  "03_umap_projected_with_unknown.pdf" = "The reference-projected UMAP, same colouring.",
  "03a_umap_projected_unknown_highlighted.pdf" = "Unknown cells on the projected UMAP.",
  "04_unknown_nearest_neighbor_identity_umap.pdf" = paste(
    "If each Unknown cell is forced to take the majority label of its 15 nearest annotated",
    "neighbours in the object's own embedding, they come out mostly monocyte and",
    "pro-monocyte."),
  "04_unknown_nearest_neighbor_identity_umap_projected.pdf" = paste(
    "The same vote in reference space instead swings to late erythroid. The two embeddings",
    "disagree, which is the tell: these cells are being placed by what they lack rather than",
    "by what they express, so their apparent identity depends on where you ask.")
)

for (f in sort(all_pdfs(unmap_dir))) {
  show_fig(f, unotes[[basename(f)]])
}
```

```{r nn-tables, results='asis'}
show_table(u_nn_nat, "Nearest annotated neighbour, native embedding", n = 8)
show_table(u_nn_proj, "Nearest annotated neighbour, projected embedding", n = 8)
```

# 4. How much the surviving labels can be trusted

**There is only one cutoff in the whole annotation, and it is the mapping-error one.**
`predict_CellTypes()` calls `knnPredict_Seurat()` with **k = 30**: each query cell is given
the majority label of its 30 nearest reference neighbours in harmonised PCA space.
`predicted_CellType_prob` is the fraction of those 30 that voted for the winning label - so
it is a vote share, not a posterior probability, and every value it can take is a multiple of
1/30. In LK2 it takes 25 distinct values, the lowest being 0.2 (6 of 30).

`predict_CellTypes()` then does exactly one thing with that number: nothing. It never
thresholds on it. The only cell it sets to NA is the cell whose `mapping_error_QC` is Fail.
So a label assigned on a 7-to-30 plurality is stored with the same standing as one assigned
unanimously, and it is up to whoever reads the object to decide whether to trust it. Nothing
downstream in this project currently filters on it either.

That makes the distribution below worth reading properly rather than skimming: it is the only
description of how firm each label is.

```{r conf, results='asis'}
show_table(t_conf, "Label confidence over cells that pass QC")
show_fig(file.path(bench_dir, "03_label_confidence.pdf"),
  paste("One bar is one neighbour changing its vote. The dashed lines are reading aids, not",
        "thresholds the tool applies: 1/30 is the floor, 0.5 is where a label stops being a",
        "true majority and becomes a plurality, 1.0 is unanimous. 34% of LK2 labels are",
        "unanimous; 8% sit below the majority line and are the ones to be careful with."))
show_fig(file.path(bench_dir, "04_label_confidence_by_celltype.pdf"),
  paste("Median confidence by broad cell type. The low end is HSC MPP, LMPP, Early Lymphoid,",
        "Cycling Progenitor and Early GMP; the high end is Monocyte, cDC, pDC, Plasma Cell",
        "and Late Erythroid."))
```

This is not a quality gradient. The populations with low confidence are the ones that are
genuinely continuous - a cell part way along the HSC-to-LMPP axis should not get a unanimous
vote, and the fact that it does not is the reference behaving correctly. It does mean that
any analysis that treats HSC MPP as a discrete population is working with a soft boundary,
which is worth remembering for the biotin split and for the surfaceome comparisons.

```{r conf-type-table, results='asis'}
if (!is.null(t_conf_ct)) {
  show_table(t_conf_ct %>% arrange(median_prob) %>% head(10),
             "Ten least confident broad cell types (>= 20 cells)")
  show_table(t_conf_ct %>% arrange(desc(median_prob)) %>% head(10),
             "Ten most confident")
}
```

# 5. What predict_CellTypes() actually does

```{r relabel, results='asis'}
show_table(t_relabel, "initial_CellType versus predicted_CellType")
```

`initial_CellType` is never NA, `predicted_CellType` is NA for exactly the QC-failed cells,
and wherever both are present they are identical for 100% of cells in both runs. So the
second call is a gate, not a second opinion: it does not revise labels, it withholds them.
That matters for anyone reading the object - `initial_CellType` is not a draft that got
improved, it is the same answer without the QC applied.

# 6. Fine versus broad labels

```{r vocab, results='asis'}
show_table(t_vocab, "Fine and broad label vocabularies")
```

This is exactly what happened to the pilot's `10_lineage_composition_by_sampleID.csv` and
`12_specific_celltype_composition_by_sampleID.csv` - the latter has an all-NA cell type
column on disk. Step 01 here is keyed on the broad column, and step 02 (the LK2 version that
became the pipeline process) already was.

# 7. The mapping-error threshold

2.5 is the BoneMarrowMap default, and it is what the standalone scripts in steps 01 and 02
ran. **The pipeline does not use it.** `nextflow/bin/annotate.R` runs at `MAD_threshold = 4`;
only `nextflow/bin/lib_early_qc_annotation.R` still has 2.5. So the figures in
`results/seurat_annotated/` are at 2.5 and anything coming out of `results_nf/` is at 4, and
the two are not interchangeable.

The score itself does not depend on the threshold - `calculate_MappingError()` stores
`mapping_error_score` per cell and then calls Pass/Fail on
`score < median(score) + MAD_threshold * mad(score)`, with median and MAD taken over the whole
object. So the sweep below is computed directly from the stored score rather than by
re-projecting. Recomputing the 2.5 call reproduces the stored `mapping_error_QC` for 100% of
cells in both runs, which is what makes the other two columns trustworthy.

```{r thresh, results='asis'}
t_sweep_run <- read_if(file.path(bench_dir, "12_MAD_threshold_sweep_by_run.csv"))
t_sweep_samp <- read_if(file.path(bench_dir, "11_MAD_threshold_sweep_by_sample.csv"))

show_table(t_sweep_run, "Rejection rate by run, at MAD_threshold 2.5, 3 and 4")

if (!is.null(t_sweep_samp)) {
  wide <- t_sweep_samp %>%
    dplyr::select(run, sample, MAD_threshold, percent_rejected) %>%
    pivot_wider(names_from = MAD_threshold, values_from = percent_rejected,
                names_prefix = "MAD ") %>%
    arrange(desc(`MAD 2.5`))
  show_table(wide, "Rejection rate per sample, worst first (percent of that sample's cells)")
}

show_fig(file.path(bench_dir, "06_MAD_threshold_sweep.pdf"),
  paste("Every sample moves in the same direction and keeps its rank - the threshold changes",
        "how much is cut, not which samples are hardest to place. APOP576-TP53 still loses",
        "11.6% at 4, more than any other sample loses at 2.5."))
```

Going from 2.5 to 4 roughly halves the rejection rate - 7.95% to 3.91% in the pilot, 5.48% to
1.75% in LK2 - and 3 sits between them. Since section 3 shows the rejected cells are shallow,
ribosome-poor and enriched for HTO negatives, the cells recovered by relaxing the threshold
are mostly ones there is no good reason to keep. That argues for 2.5 rather than the 4 the
pipeline currently runs, but it has not been changed, so the discrepancy stands and is worth
resolving deliberately rather than by drift.

# 8. Projection output

The projected UMAP for both runs, coloured by `predicted_CellType_Broad` in the project
palette. The versions in `results/seurat_annotated/` are coloured by the 53-level fine label
in ggplot defaults and exist for one run at a time, which is why step 04a redraws them.

```{r projection, results='asis'}
show_fig(file.path(bench_dir, "07_projected_umap_broad_both_runs.pdf"),
  paste("LK1 pilot and LK2 side by side. Same reference space, same palette, so the",
        "compositional difference between the two runs can be read directly off the map."))
```

# 9. What was run

```{r steps, results='asis'}
steps <- tibble::tribble(
  ~Step, ~Script, ~What,
  "00", "00.celltype_palette.R",
  "The bespoke broad-cell-type palette, lineage vocabulary and the check_palette() guard, defined once and sourced by steps 01-03. Run directly, it redraws the reference UMAPs.",
  "01", "01.annotate_LK1_pilot.R",
  "BoneMarrowMap projection of the LK1 pilot, MAD 2.5.",
  "02", "02.annotate_LK2.R",
  "The same projection on LK2. This is the version that became the pipeline's 08_annotate process.",
  "03", "03.check_unannotated_cells.R",
  "Re-projects the LK2 pipeline checkpoint at MAD 4 and characterises the cells that still fail - QC covariates, donor, HTO class, native cluster, nearest annotated neighbour.",
  "04", "04.annotation_benchmark_summary.R",
  "Scores the annotation from the metadata tables alone: reject rates, rejected-cell profile, label confidence, initial-versus-final label, fine-versus-broad vocabulary, the MAD threshold sweep.",
  "04a", "04a.projected_umaps.R",
  "Redraws the projected UMAP for both runs in the broad palette. Separate from step 04 because it is the only step that loads the ~450 MB projected objects.",
  "05", "05.celltype_annotation_report.Rmd",
  "This report."
)
show_table(steps)
```

Steps 01 and 02 are collated from `backup/10.annotate.R`, `backup/10a.annotate.R`,
`backup/10b.annotate_DSB.R` and `8.annotate.R`. `10.annotate.R` is superseded - none of its
output names survive in `results/`. `10a` and `10b` are the same script run twice with
different ADT assays, which is now the step 01 argument. Three fixes went in on the way:

- `10a` had a `marker_map` / `plot_df` block sitting above the code that creates the objects
  it references, so it could only ever have been run interactively out of order. Removed.
- `10a.heatmap_cell_type_all.pdf` is 0 bytes on disk because scaling a zero-variance row
  gives NaN and `pheatmap` will not draw it. Fixed in both steps.
- the numbat export loops at the tails of `10a` and `8.annotate.R` belong to the CNV work,
  not to annotation. Dropped from both.
- every script carried its own copy of the palette, and the copies had drifted. All of them
  listed `GMP` - a label the reference never emits, since it splits that population into
  Early GMP and Late GMP - and none had an entry for `Stromal`, which the reference does
  emit. Because the plots do `factor(CellType_Broad, levels = names(celltype_cols))`, every
  stromal cell was being set to NA and dropped at the plotting step. That is why stromal
  cells looked unannotated: they were annotated, they were being discarded when drawn. The
  palette now lives in step 00 with `Stromal` in grey, and `check_palette()` raises a warning
  instead of silently dropping anything it does not recognise.

# 10. Data locations

```{r locations, results='asis'}
locs <- tibble::tribble(
  ~What, ~Where,
  "BoneMarrowMap reference + uwot model", "annotation/BoneMarrowMap_SymphonyReference.rds, annotation/BoneMarrowMap_uwot_model.uwot",
  "Pilot input (CITE/DSB object)", sprintf("results/cite_qc_then_dsb/%s/LK1-GEX/seurat_demux_ADT_CLR_DSB_QC.rds", pilot),
  "LK2 input (CITE/DSB object)", sprintf("results/cite_seq_dsb/%s/LK2-GEX/seurat_demux_ADT_CLR_DSB_QC.rds", run),
  "LK2 pipeline checkpoint (step 03 input)", sprintf("results_nf/%s/rds/07_cite_qc/LK2-GEX/rds/seurat_demux_ADT_CLR_DSB_QC.rds", run),
  "Pilot annotation outputs", sprintf("results/seurat_annotated/%s", pilot),
  "LK2 annotation outputs", sprintf("results/seurat_annotated/%s", run),
  "Unannotated-cell investigation", sprintf("results/seurat_annotated/%s/LK2-GEX_results_nf_unannotated_check", run),
  "Benchmark tables and figures (step 04)", "results/benchmarking/celltype_annotation"
)
show_table(locs)
```

```{r tools, results='asis'}
tools <- tibble::tribble(
  ~Tool, ~Version, ~Use,
  "BoneMarrowMap", "0.2.0", "Zeng et al. normal bone marrow Symphony reference and helpers",
  "symphony", "0.1.2", "reference mapping",
  "uwot", "0.2.4", "UMAP projection into reference coordinates",
  "Seurat", "5.4.0", "object handling, DimPlot/FeaturePlot",
  "class", "7.3.23", "kNN vote for the Unknown-cell identity assignment",
  "pheatmap", "1.0.13", "composition heatmaps"
)
show_table(tools, "Tools")
```

# 11. Caveats

- steps 01 and 02 are the collated, cleaned versions of the scripts that produced the
  original annotation, and were not re-executed - the projections they wrote in May and June
  are what everything here is computed from. Steps 00, 03, 04 and 04a were run fresh, which
  is what `Analysis finished` reflects. Step 03 reused its saved annotated object rather than
  re-projecting, so only its figures are new, not its numbers
- the pilot and LK2 numbers come from two different input objects and two different upstream
  CITE QC runs, so the difference in reject rate between runs is not a controlled comparison.
  The per-sample ordering within each run is, and so is the threshold sweep, which moves the
  same cells against the same score
- step 03 runs on the pipeline checkpoint at `MAD_threshold = 4`, not on the step 02 object at
  2.5. Its 1.35% Unknown and step 02's 5.48% Fail are therefore both correct and not directly
  comparable - the sweep in section 7 is the like-for-like version, since it recomputes both
  thresholds against the same cells
- the ADT/CITE marker projections that steps 01 and 02 also write are not shown here. The
  step 01 `ADT_ASSAY` argument still selects between `ADT_CLR` and `CITE_DSB` for them, and
  both were run; the figures are on disk under `projectionFigures/`
- APOP576-TP53's 30% reject rate has not been chased down beyond noting it. It is the sample
  to look at first if that number matters
- pseudotime is inherited from the reference, not computed on our cells, so it should be read
  as "where in the normal trajectory does this cell project", not as a trajectory of the
  leukaemia
